--- 
title: "Reproducible Research in R"
subtitle: "An introductory workshop on modern data analyses and workflows"
author: "Luke W. Johnston"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [resources/refs.bib, resources/packages.bib]
link-citations: yes
description: "This is a 3-day workshop introducing modern data analysis tools and workflows for researchers."
---

# Syllabus

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), here::here('resources/packages.bib'))
```

> NOTE: This content is still being updated.

Reproducibility and open scientific practices are increasingly demanded of
scientists and researchers. Training on how to apply these practices in data
analysis has not kept up with demand. With this course, we hope to begin meeting
that demand. Using a very practical approach bases mostly on code-along sessions
(instructor and learner coding together), the course will:

1. Explain what an open and reproducible data analysis workflow is, what it
looks like, and why it is important.
1. Explain and demonstrate why R is rapidly becoming the standard program of
choice for doing modern data analysis in science.
1. Demonstrate and apply collaborative tools and techniques when working in team
settings (including working with your future self).
1. Show and apply the fundamental tools and skills for conducting a
reproducible and modern analysis for a research project.
1. Show where to go to get help and to continue learning modern data analysis
skills.

We'll be addressing the following questions:

- What is [R], why should I use it, and how do I use it?
- What does a modern data analysis setup and workflow look like?
- What is reproducibility and how is it different from replicability?
- How can I ensure my data analysis project is reproducible?
- How can I import and work with my data in R?
- How can I visualize my data and make publication-quality figures?
- Why should I and how can I keep track of changes to my analysis files?
- How can I write reports to document, describe, and present analyses in a
reproducible way?

By the end of the course, participants will have a basic level of proficiency in
using the [R] statistical computing language, enabling them to improve their data
and code literacy, and to conduct a modern and reproducible data analysis. The
course will place particular emphasis on research in diabetes and metabolism; it
will be taught by [instructors](#instructors) working in this field and it will
use relevant examples where possible.

[R]: https://www.r-project.org/

## Is this course for you?

To help manage expectations and develop the material for this course, we make a
few assumptions about *who you are* as a participant in the course:

- You are a researcher, likely working in the biomedical field (ranging from
experimental to epidemiology).
- You currently or will soon do some quantitative data analysis.
- You: 
    - know *nothing* or *little* about R (or computing in general); 
    - *haven't* used coding programs for doing data analysis (e.g. used SPSS); 
    - *have* used coding programs before (e.g. used SAS or Stata), but not R;
    - know how to use R, but *haven't* used the [tidyverse] or [RStudio].

[tidyverse]: https://www.tidyverse.org/
[RStudio]: https://rstudio.com/

While we have these assumptions to help focus the content of the course, if you
have an interest in learning R but don't fit any of the above assumptions, *you
are still welcome to attend the course*!! We don't turn anyone away if we can.

In addition to the assumptions, we also have a fairly focused scope for teaching
and expectations for learning. So this may also help you decide if this course
is for you.

- We *do* teach how to use R, starting from the very basics and targeted to beginners.
- We *do not* teach statistics (these are already covered by most university curriculums).
- We *do* teach from a team science, reproducible research, and open scientific
perspective (i.e. by including a collaborative group project that uses a transparent
and reproducible analysis workflow).
- We *do* teach using practical, applied, and hands-on lessons and exercises, 
with a few short lectures that introduce a topic.

To further develop your R skills and knowledge, we are having an advanced R
course from September 8-9, 2020 that will build off of this course (though it
isn't dependent on this course). Keep on eye on DDA announcements to register
and learn more about it!

## General schedule

The workshop is structured as a series of participatory live-coding sessions
(instructor and learner coding together) interspersed with hands-on exercises,
using either a practice dataset or the participants' own datasets. Some short 
lectures will be given throughout.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
options(knitr.kable.NA = "")
schedule <- readr::read_csv("data/schedule.csv")
knitr::kable(schedule, align = "l")
```

Notes: Describe assignments and form groups during introduction.
Get them in line from "complete beginner to computing" to "know a little bit".
Then assign numbers for groups then. Structure room into group tables, but get 
them to sit wherever but not get comfortable when they sit down (since they will
move).

## Instructors

Recommended number of instructors for a 25-35 class size is about 3-6
instructors (~1 instructors for every 1-2 sessions). If there are more instructors,
then less helpers are needed. The recommended combined total number of helpers
and instructors is about 6 for a smooth workflow experience for the learners.

## Recommended resources

- [R for Data Science](http://r4ds.had.co.nz/): Excellent open and online resource for using R for data analysis and data science.
- [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/): Excellent online resource for using ggplot2 and R graphics.
- [RStudio cheat sheets](https://www.rstudio.com/resources/cheatsheets/): Great quick reference.
