# Resources

## For this course

### Potential exercise solutions

These are *potential* exercise solutions of R code only. This are mostly intended
as a resource for after the class, not during it.

```{r all-exercise-solutions, results="asis", echo=FALSE, eval=TRUE, message=FALSE}
library(magrittr)
source(here::here("R/extract-solutions.R"))
chapter_files <- fs::dir_ls(regexp = "^0[1-5]-.*.Rmd$")
solutions <- purrr::map(
    chapter_files,
    extract_solution_chunks
)
unlist(solutions) %>% 
    cat(sep = "\n")
```


## For learning

**Free online books**:

- [R for Data Science](https://r4ds.had.co.nz/): 
Excellent open and online resource for using R for data analysis and data science.
- [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/): 
Excellent online resource for using ggplot2 and R graphics.
Book mostly focused on concepts and theory of how to visualize, 
rather than the practicalities (i.e. no coding involved).
- [ModernDive: Statistical Inference via Data Science](https://moderndive.com/): 
Great book on using statistics and data science methods in R
- [Happy Git and GitHub for the useR](https://happygitwithr.com/) (**highly recommended**):
Specifically useful is the chapter on [Daily Workflows](https://happygitwithr.com/repeated-amend.html) using Git
- [Data Visualization: A practical introduction](https://socviz.co/index.html#preface):
A book that goes into practical as well as conceptual detail on how 
and why to make certain graphs, given your data.
- [Online book for R Markdown](https://bookdown.org/yihui/rmarkdown/):
The go-to reference for learning and using R Markdown
- [Course material for a statistics class](https://stat545.com/):
Excellent course material for teaching statistics and R.
    

**Quick references**:

- [RStudio cheatsheets](https://www.rstudio.com/resources/cheatsheets/):
Multiple, high-quality cheatsheets you can print off to use as a handy reference
- [Tidyverse style guide](https://style.tidyverse.org/):
To learn about how to write well styled code in R

**Articles**:

- [Good enough practicies in scientific computing](https://doi.org/10.1371/journal.pcbi.1005510):
An article listing and describing some practices to use when writing code
- [Best practices in scientific computing](https://doi.org/10.1371/journal.pbio.1001745)
- Case study of reproducible methodds in Bioinformatics: [@Kim2018a]

**General sites**

- [Case studies](https://www.practicereproducibleresearch.org/) and lessons for doing reproducibility
- [Organizing R Source Code](https://www.r-bloggers.com/r-best-practices-r-you-writing-the-r-way/)
- [Hands-on tutorial for learning Git, in a web-based terminal](https://try.github.io/levels/1/challenges/1)
- [Simpler, first-steps guide to using Git](https://rogerdudler.github.io/git-guide/)
- [RStudio tutorial on using R Markdown](https://rmarkdown.rstudio.com/lesson-1.html)
- [Markdown syntax guide](https://rmarkdown.rstudio.com/authoring_basics.html)
- [Pandoc Markdown Manual](https://pandoc.org/MANUAL.html#pandocs-markdown) (R
Markdown uses [pandoc])
- [Adding citations in R Markdown](https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

**Interactive sites or resources for hands-on learning**:

- [Interactive tutorials for using R, within R](https://swirlstats.com/)
- [RStudio's Learning Primers](https://rstudio.cloud/learn/primers)

**Videos**:

- Video on using [Git in RStudio]

[Git in RStudio]: https://rstudio.com/resources/webinars/managing-part-2-github-and-rstudio/

**Getting help**:
    
- [StackOverflow for tidyr](https://stackoverflow.com/questions/tagged/tidyr)
- [StackOverflow for dplyr](https://stackoverflow.com/questions/tagged/dplyr)
- [StackOverflow for ggplot2](https://stackoverflow.com/questions/tagged/ggplot2?sort=frequent&pageSize=50)
- Tip: Combine auto-completion with `::` to find new functions 
and documentation on the functions 
(e.g. try typing `base::` and then hitting Tab to show a list of all functions found in base R)

```{r}
knitr::kable(r3::useful_packages_list, caption = "Useful and common packages to use in data analysis.")
```

