---
title: "What can you do next after this course?"
subtitle: "Doing it and learning from other researchers"
---

layout: true

```{r setup, include=FALSE}
source(here::here("R/slide-setup.R"))
```

```{r}
xaringanExtra::use_progress_bar(color = progress_bar_colour, location = "top")
```

---

## What else can you do?

- Find or start building a community of people using R

- Start reviewing each others code

- Start sharing your code

- Think about how your code will be reproducible

- Teach or help out with R courses `r emo::ji("wink")` 

- Join local initiatives that follow these guidelines

???

The other things you can start doing is find or start building a community of
people who also use R or are doing reproducibility or any other computational
work. Use them as support and help and also give back too.

Start doing code reviews in your research group. Code review would be where you
look over each others code, check that it works, check that it makes sense, 
that it's readable and understandable. The nice thing with doing code reviews
is that it dispels the mystery around code and about criticising it and trying
to improve it. We review manuscripts, why not code?
I personally though have had a really hard time getting groups I've been part of
now and in the past to do this, but baby steps.

You can teach. Teach others. Use these teaching materials. Or get involved with
this course next year. Or now. Actually, several participants from the beginner
course are or will soon be helping to improve the material for next year. There
are also so many other things you can get involved in, aside from this course.
Let me know if you're interested!
