# Importing data into R {#sec-import-data}

## Learning objectives

This session’s **overall learning outcome** is to:

1.  Identify the different ways to import data by relying on the file
    extension, describing and identifying the "tidiness" of the data,
    and using R to import the data.

**Specific objectives** are to:

1.  Describe different ways of importing data into R and identify which
    method to use based on the file extension, then use one of these
    methods to import a dataset.
2.  Explain how to ensure your project is self-contained by using
    relative file paths within the R project and use the `here::here()`
    function to make these paths.
3.  Use a tool called "auto-completion" to more quickly type out objects
    in R.
4.  Use options in Quarto Markdown's code chunks to control what gets
    shown from the output of the code chunk when generating the output
    document.
5.  Describe the difference between "messy" and "tidy" data and explain
    why you should aim to make data tidier.
6.  Identify some approaches to help resolve problems when you encounter
    errors or warnings in R.

## :book: Reading task: Different ways of importing data

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Very briefly go over this section again, mainly emphasizing that how you
import your data depends on what data file format it is in.
:::

**Time: \~4 minutes**

Before you can do any type of data analysis, you first need to import
the data into R. There are several ways to import in a dataset, which
are listed below. *Don't run these code, just read for now*.

1.  Using the RStudio menu
    `File -> Import Dataset -> From Text/Excel/SPSS/SAS/Stata`
    (depending on your file type you want to import). This approach will
    also generate the code for you to use in the future, which you
    should copy and paste to your Quarto document or R script.

2.  If the file is a `.csv` file, use `readr::read_csv()` to import the
    dataset. This is what we will be doing shortly.

3.  If the dataset is a `.rda` file, use `load()`:

    ``` r
    load(here::here("data/dataset_name.rda"))    
    ```

    This loads the dataset into your R session so that you can use it
    again.

4.  For SAS, SPSS, or Stata files, you can use the package
    [haven](https://haven.tidyverse.org/) to import those types of data
    files into R.

{{< text_snippet sticky_up >}}

## Importing a CSV file

Now is the time to import some data into our R project. But we're
missing something: the data itself! We'll download an [open
dataset](https://zenodo.org/records/4989220) from Zenodo
[@Eliasson2018], which is a online open science archive for various
research outputs, like datasets, preprints, software, protocols, and
teaching material. The data is from a study looking at the impact of a
meal on insulin levels in different groups.

We want to download the data to our R project. First, make sure you are
in your `LearningR` R project. Then copy and paste the code below into
the **Console** and run it.

```{r}
#| filename: "Console"
#| eval: false
download.file(
  url = "https://zenodo.org/records/4989220/files/Eliasson_data2.csv?download=1", 
  destfile = here::here("data/post-meal-insulin.csv")
)
```

This function downloads the file from the Zenodo website and then saves
it to the `data/` folder in your R project with the name
`post-meal-insulin.csv`. We are using the Console and **not** the Quarto
document because we don't want to run this code every time we render the
document. There is the `here::here()` function in the code, which we
will explain in a bit.

Before we import the data into the R session, it's useful to have a
quick manual look at data to see if there are any things we should be
aware of or consider when we do import it. So go to the File pane, go to
the `data/` folder, click the `post-meal-insulin.csv` file, then click
the "View file" option to open it up in RStudio. When we look at it
directly, we see something that looks like:

```{r}
#| echo: false
readr::read_lines(here::here("data/post-meal-insulin.csv")) |>
  stringr::str_trunc(60) |>
  head() |>
  cat(sep = "\n")
```

You'll notice that the data is separated by semicolons (`;`) and not
commas (`,`). This is a common issue when working with data from Europe,
as they often use semicolons as the delimiter since commas are often
used for decimal places in numbers. For data actually separated by
commas in *comma separated value* (CSV) files, we would use the
`read_csv()` function in the `{tidyverse}` package. But for data
separated by semicolons, we use the `read_csv2()` function instead.

So, in the `setup` code chunk, write out the code to import (read) the
data into R.

```{r setup}
#| echo: fenced
library(tidyverse) 
post_meal_insulin_data <- read_csv2(here::here("data/post-meal-insulin.csv"))
```

In the code chunk above, we are using the `read_csv2()` function that is
technically from the `{readr}` package, which is loaded in with
`{tidyverse}`, to import the dataset. We're also again using the
`here::here()` function, but there is a short reading task below to
explain it, so we won't cover this just yet. Let's run this code by
having our cursor of the line and using {{< var keybind.run-code >}}.
There will be some output (like shown above this paragraph). This output
message informs you what R is doing when reading in the data and gives
some basic details about the data.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Briefly explain the output of the `read_csv2()` function to them.
:::

Let's print out the data into our Quarto document to see what it looks
like. At the bottom of your `docs/learning.qmd` file, create a new level
2 header called "Showing the data" and insert a new code chunk with
{{< var keybind.chunk >}}, which should look like the following:

```` markdown

## Showing the data

```{{r}}

```
````

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Walk through how to use auto-completion.
:::

To more quickly type out objects in R, use "tab-completion" or
"auto-completion" to finish an object name for you. Normally RStudio
will start auto-completing for you as you type code, but you can
manually trigger auto-completion with `Tab`. As you type out an object
name, hit the `Tab` key to see a list of objects available. RStudio will
not only list the objects, but also shows the possible options and
potential help associated with the object. Let's do that for the new
object we created by starting to type:

```{r}
#| filename: "docs/learning.qmd"
#| eval: false
post_
```

Then hit tab. You should see a menu pop up with a list of potential
matches. Hit tab again to finish with `post_meal_insulin_data`. This
simple tool can save so much time and can prevent spelling mistakes.
Let's finish adding the new data frame object into the code chunk.

```` markdown

## Showing the data

```{{r}}
post_meal_insulin_data
```
````

Run the code using {{< var keybind.run-code >}} to see the output. This
way of showing the data can be useful to see bits of the data. Another
way of getting a bit better view of the data is using the `glimpse()`
function. So, let's use `glimpse()` below the code we just write. Don't
forget to use auto-completion!

```{r}
#| filename: "docs/learning.qmd"
post_meal_insulin_data 
glimpse(post_meal_insulin_data)
```

Run this line of code with {{< var keybind.run-code >}} to see the
output. Using `glimpse()` is a great way to see the structure of the
data, including the column names, the data type of each column, and the
first few rows of the data. Let's render the document with
{{< var keybind.render >}} to see what the output HTML document looks
like.

Before moving on to the reading task, let's add the change's we've made
to the Git history with {{< var keybind.commit >}}. If you work with
personally sensitive (health) data, you would not commit the data file
so that we don't accidentally put it publicly on GitHub. Or if your data
is really large, we don't normally want to keep large files on GitHub
since that isn't the best place to store those types of files. But
because the dataset is already an open dataset and because it is small,
for this course we will save it to the Git history to get practice using
Git. We also normally don't commit the generated `.html` files, but for
this course we will commit them to the Git history so that you get
practice with Git. Once done, read over the next section.

## :book: Reading task: Relative paths in the R project

**Time: \~5 minutes**

The `here()` function from the `{here}` package, usually referred to as
"here here" and written with `here::here()` so that we don't have to
load the package, helps to make it easier to manage *file paths* within
an R Project.

So, what is a file path and why is this `{here}` package necessary? A
file path is the list of folders a file is found in. For instance, your
CV may be found in `/Users/Documents/personal_things/CV.docx`. The
problem with file paths when running code (like with R) is that when you
run a script interactively (e.g. what we do in class and normally), the
file path and "working directory" (the R session) are located at the
Project level (where the `.Rproj` file is found). You can see the
working directory by looking at the top of the RStudio Console.

But! When you render a Quarto document, run an R script, or run the code
in a non-interactive way, the R code may likely run *in the folder it is
saved in*, e.g. in the `R/` folder. So your file path
`data/post-meal-insulin.csv` won't work because there isn't a folder
called `data/` in the `R/` folder.

```         
LearningR <-- R Project working directory starts here.
├── R
│   └── README.md
├── data
│   └── README.md
├── data-raw
│   └── README.md
├── docs
│   ├── learning.qmd <-- Working directory when running not interactively.
│   └── README.md
├── .gitignore
├── DESCRIPTION
├── LearningR.Rproj <-- here() moves file path to start in this file's folder.
├── README.md
└── TODO.md
```

Often people use the function `setwd()` in scripts, but this is *never*
a good idea since using it makes your code *runnable only on your
computer*. Which means, it is not reproducible! We use the `here()`
function to tell R to go to the project folder (where the `.Rproj` file
is found) and then use that file path. This simple function can make
your work more reproducible and easier for you and others to use later
on.

{{< text_snippet sticky_up >}}

## Setting code chunk options

One thing you may have noticed is that when we render the document to
HTML, there is a bunch of extra output, especially below the `setup`
code chunk. You probably don't want this text in your generated
document, so we will add a [chunk
option](https://quarto.org/docs/reference/cells/cells-knitr.html) to
remove this message.

Chunk options are used to change how code chunks work. When adding them
inside the code chunk, they always need to start with `#|`. In this
case, we don't want the messages or warnings when we load `{tidyverse}`
or read in the dataset. The options to remove those messages and
warnings are `#| message: false` and `#| warning: false`, so let's add
those to the `setup` code chunk.

```{r setup}
#| echo: fenced
#| message: false
#| warning: false
library(tidyverse)
post_meal_insulin_data <- read_csv2(here::here("data/post-meal-insulin.csv"))
```

Let's render the HTML document again with {{< var keybind.render >}} and
see what changes.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Briefly show these other options below to them on the screen and explain
it, but don't get them to read it.
:::

Other common options are:

-   `include`: Whether to include all the code, code output, messages,
    and warnings in the rendered output document. Default is `true`. Use
    `false` to hide everything but still run the code.
-   `echo`: To show the code. Default value is `true`. Use `false` to
    hide.
-   `results`: To show the output. Default is `markup`. Use `hide` to
    hide or `asis` as regular text (not inside a code block).
-   `eval`: To evaluate (run) the R code in the chunk. Default value is
    `true`, while `false` does not run the code.

These options all work on the individual code chunk. If you want to set
an option to all the code chunks (e.g. to hide all the code but keep the
output), you can use Quarto's
[`execute`](https://quarto.org/docs/reference/formats/html.html#execution)
options. These options are added to the YAML header and will apply the
settings to everything in the document. We *won't* do this in this
session, but here is what it looks like:

``` {.yaml filename="docs/learning.qmd"}
---
title: "Reproducible documents"
author: "Your Name"
format: html
execute:
  echo: false
  warning: false
  message: false
---
```

Before moving on, let's commit the changes we've made to the Git history
with {{< var keybind.commit >}}.

## :book: Reading task: Working with messy and tidy data

::: {.callout-note collapse="true"}
## Instructor note

For the reading section, emphasize the characteristics of a "tidy"
dataset.
:::

**Time: \~12 minutes**

The concept of "tidy" data was popularized in an
[article](https://www.jstatsoft.org/v59/i10/paper) [@Wickham2014] by
Hadley Wickham and described in more detail in the [Tidy Data
chapter](https://r4ds.hadley.nz/data-tidy.html) of the *R for Data
Science* online book. Before we continue with tidy data, we need to
cover something that is related to the concept of "tidy" and that has
already come up in this course: the `{tidyverse}`. The `{tidyverse}` is
an ecosystem of R packages that are designed to work well together, that
all follow a strong "[design philosophy](https://design.tidyverse.org/)"
and common [style guide](https://style.tidyverse.org/). This makes
combining these packages in the tidyverse much easier. We teach the
tidyverse because of these reasons.

Part of the "tidy" part of tidyverse revolves around tidy data. A tidy
dataset is when:

-   Each variable has its own column (e.g. "Body Weight").
-   Each observation has its own row (e.g. "Person").
-   Each value has its own cell (e.g. "Body weight for a person at a
    specific date").

Take a look at the example "tidy" and "messy" data frames (also called
"tibbles" in the tidyverse) below. Think about why each may be
considered "tidy" or "messy". What do you notice between the tidy
versions and the messier versions?

```{r}
# Datasets come from tidyr
# Tidy:
table1

# Partly tidy:
table2

# Messier:
table3

# Messy:
table4a

# Messy:
table4b
```

The "most" tidy version is `table1` as each column describes their
values (e.g. population is population size), each row is unique (e.g.
first row is for values from Afghanistan from 1999), and each cell is an
explicit value representative of its column and row.

`table2` is a "long" version of `table1` so it is partly "tidy", but it
doesn't satisfy the rule that each variable has a column, since `count`
represents both cases and population size.

On the other hand, `table3` is messy because the `rate` column values
are a composite of two other column values (cases and population), when
it should be a single number (a percent). Both `table4a` and `table4b`
have columns with ambiguous values inside. For example, you can't tell
from the data what the values in the `1999` column contain.

Tidy data has a few notable benefits:

1.  Time spent preparing your data to be tidy from the beginning can
    save days of added work and frustration in the long run.
2.  "Tidy data" is a conceptual framework that allows you to easily
    build off and wrangle (i.e. "manipulate", "clean up", "manage") data
    in simpler and easy-to-interpret ways, especially when used within
    the framework of the tidyverse.
3.  Tidier data is easier to visualize and "map" onto different axises
    of the plot, like the x and y axises.

The concept of tidy data also gives rise to "tidy code" for visualizing
and wrangling. By using "verbs" (R functions) and chaining them together
in "sentences" (in a sequential pipeline), you can construct meaningful
and readable code that describes in plainer English what you are doing
to the data. This is one simple way that you can enhance the
reproducibility of your code. There are other ways to make your code
tidier and more readable, all of which we will cover in this course.

-   Document what you did to your data and why you did it, to help you
    remember later on by using comments with `#` hashes as well as using
    Markdown text to describe things that can't be easily explained by
    the code, for example, the "why" behind the code.
-   Keep the code simple: Don't be clever nor overly concise, always be
    clear even if it means more code. Clear code is easier to understand
    than clever and sometimes overly-complex or -concise code.

Whether working with either messy or tidy data, there are a few
principles to follow:

-   You should always save your original raw dataset in the `data-raw/`
    folder.
    -   *Note*: Whether or not you save data to `data-raw/` depends on
        how you collected the data and how many collaborators are on
        your team. You may end up storing and processing the data in
        another folder as a project of its own.
-   **Never** edit your raw data directly. If you have to edit it to fix
    mistakes, have some version control of it so you can track and see
    what has been changed.
-   Only work with your raw data using R code. *As much as possible or
    is reasonable, don't manually edit it*. Manual editing doesn't leave
    a history of what you've done to it, so you can't go back and see
    what you've done. Always keep a history of any changes you've made
    to the data, preferably by using R code.
-   Save the edited data as another dataset and store it in the `data/`
    folder.

We are saving to `data/` because the dataset is already collected and
published, so the original raw data can always be downloaded again.

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: How tidy is our data?

**Time: \~10**

Now that we've learned about the concept of "tidy" data, let's discuss
how tidy our data is. At a glance, there are approximately 5 things in
this data that are mildly untidy or could be improved. Can you identify
them?

1.  On your own, take 2 minutes to look over the data and try to find
    what could be improved.
2.  With your neighbour, take 5 minutes and discuss what you found and
    see if you have overlap in agreement or if you found different
    things.
3.  As a group, we will take 3 minutes and we will share what you all
    discussed and then go over the 5 things.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

The 6 main things are:

1.  Column names are not tidy or standardised. For instance, there are
    `.` in the names and some use abbreviations.
2.  Some column names also contain data. For instance, all the columns
    with numbers at the end are time points, like `Insulin1` is insulin
    measured at 1 minute after eating the meal. This data is in the wide
    format, when it should be in the long format. We'll cover long and
    wide in @sec-wrangling.
3.  The data could actually be split into two or three datasets, one in
    the long format for the measurements taken at regular minute
    intervals, one for the measurements taken at 30 minute intervals
    (the OGTT).
4.  There are two ID columns, `id` and `OFS.ID`.
5.  The `Group` column has abbreviations as `FDR` and `CTR` instead of
    the full names.
:::

## :book: Reading task: Encountering problems and finding help

::: {.callout-note collapse="true"}
## Instructor note

Briefly go over this section with them, especially emphasize "Restart
R", reading the error or warning message, and checking for missing
commas, brackets or misspelled words.
:::

**Time: \~10 minutes**

![A common and frequent experience when working in R. Artwork by
[\@allison_horst](https://github.com/allisonhorst/stats-illustrations).](../images/art-debugging.jpg){#fig-art-debugging
width="90%"}

You *will* encounter problems and errors when working with R, and you
will encounter them *all the time*. In fact, a large amount of your time
in R will be spent figuring out solutions to these errors ("debugging").

RStudio has many cheat sheets of its own that can help you in your
learning journey, which you can find with the Command Palette
({{< var keybind.palette >}}, then type "cheat sheet"). However, even
with these cheat sheets, you will still encounter other problems like
errors or warnings.

Error messages will appear in red text in your Console and will start
with the word "Error:". Warning messages are also (usually) in red text,
but are often either harmless or informative, so make sure to *read* the
message first and see if it says "Error" or not. Here are some initial
steps to take when you encounter an error:

1.  **First, try to stay calm**; problems happen to everyone, no matter
    their skill level.
2.  **Read through the error or warning message and see if you can
    understand what R is telling you**. Some common error messages
    include:
    -   *"Could not find function"*: Usually means that you have
        misspelled the function or an R package has not loaded properly.
    -   *"Object not found"*: Usually means that you have not
        initialized (created) the object or the object is initialized
        but empty.
    -   *"Error in..."*: Usually means that you are referring to an
        object that doesn't exist.
    -   *"Unexpected symbol in..."*: Usually means that you misspelled a
        variable or object name, so R can't find it.
3.  **Go over the code again and carefully check for any mistakes:**
    -   Missing commas or pipes?
    -   Missing end brackets like `]`, `)`, or `}`?
    -   Capitalized something that shouldn't be capitalized?
    -   Object or column name misspelled?
    -   Forgot to load your data before working on it?
    -   Forgot to load or re-load your packages? Packages are
        automatically unloaded when you exit RStudio and R. So you need
        to load them each new session with the `library()` function.
4.  **Go back to the start of the code and run each line one at a
    time**, to see where the problem occurs. You will get an opportunity
    to practice this later, once you are working with bigger chunks of
    code.

If you still can't find the problem, here are some other steps to take:

1.  **Restart the R session with** {{< var keybind.restart-r >}} **or
    with the menu item `Session -> Restart R`**. Then load your packages
    (and data if needed) and run the code from the beginning, tracking
    which objects get created, and if the proper object name is used
    later on.

2.  (Rarely need to do) **Close/re-open RStudio and try again.**

3.  **Use `help()` or `?` to access built-in documentation about a
    function or package.** You may be using the function incorrectly, so
    find out more about the function by looking at the built-in
    documentation. The documentation will open up in the "Help" pane of
    RStudio (bottom right-hand corner). **Try it out:** Enter either of
    the following commands into the Console and run it (hit `Enter`).

    ``` {.r filename="Console"}
    ?colnames
    help(colnames)
    ```

    Sometimes, this documentation can be hard to read and seem overly
    complex for a beginner. You can also try finding the website for the
    package you are having trouble with, as they often have guides that
    are a little easier to understand. The
    [tidyverse](https://www.tidyverse.org/packages/) packages all have
    amazing documentation that you can use to help you with problems you
    may have.

4.  **Consider explaining the problem out loud to a colleague or
    friend.** (or to yourself!) You might find that, in verbally going
    through the problem and explaining it, you will likely come up with
    the solution yourself.

5.  **Take a break and come back to it later!**

6.  **Google it.** Chances are that someone has already encountered that
    error and has asked about it online. In fact, those who are
    "experts" in coding languages like R are experts largely because of
    their skill in knowing the right words or terms or questions to ask
    Google. Usually googling the error message will be enough to find
    the answer, but sometimes you'll need to include "R" or "rstats" and
    the relevant package or function as a keyword in your search.

7.  If all else fails, you can always turn to the trusty **online R
    community**. Check
    [StackOverflow](https://stackoverflow.com/questions/tagged/r), a
    coding-related question and answer website, to see whether your
    issue has already been asked and solved by others. If it hasn't and
    you are considering submitting a question, make sure to read the
    posting guides beforehand to ensure that you are asking the question
    in a helpful way.

*Final words:* It is important to always work towards writing "better"
and "neater" code, as this can make it easier to break down pieces of
code and troubleshoot problems. Knowing how to do that takes some
experience, that you can only get by practicing more coding!

{{< text_snippet sticky_up >}}

## Summary

-   Use `read_csv()` or `read_csv2()` to import data from a `.csv`
    (comma or tab separated) file.
-   Use `here::here()` to make file paths easier to manage in your R
    project.
-   Use `glimpse()` to get an overview of the data.
-   Use the Quarto code chunk options like `echo`, `message`, or
    `warning` to control what is shown in the output of the code chunk.
-   Use `?` to get help on an R object. Use the cheat sheets to help
    guide you in learning R.
-   Use tab auto-completion when writing code.
-   With tidy data, each variable has its own column, each observation
    has its own row, and each value has its own cell.
-   Never edit raw data. Instead, use R code to make changes and clean
    up the raw data, rather than manually editing the dataset.

{{< include /includes/_code-appendix.qmd >}}
