# Analytically reproducible documents {#sec-reproducible-documents}

## Learning objectives

This session's **overall learning outcome** is to:

1.  Describe what a reproducible document is, explain why it is
    important, and use it to write R code to create an output document.

**Specific objectives** are to:

1.  Describe what Quarto is and how it helps with reproducibility, and
    why it can save you time and effort.
2.  Write and use R code within a document, so that it will
    automatically insert the R output into the final document.
3.  Identify what some Markdown syntax does to the formatting of the
    generated HTML file and write some Markdown to make some simple
    formatting of your document.
4.  Define some basic objects and data types in R and write some simple
    R code within the Quarto document.

## :book: Reading task: Why and how to be reproducible?

**Time: \~10 minutes**

Both reproducibility and replicability are cornerstones for doing
rigorous and sound science. Unfortunately, reproducibility in science is
near to non-existent. However, being reproducible isn't just about doing
better science. It can also mean that:

1.  You are much more efficient and productive, as less time is spent
    between coding and transferring your results to a document. No need
    to copy and paste!
2.  You can be confident in your results, since what you report and show
    as figures or tables will be exactly what you get from your
    analysis. Again, no copying and pasting required!

![Have a more reproducible workflow by using Quarto. Artwork by
\@allison_horst.](/images/art-reproducibility.png){width="70%"}

One common way of being more reproducible when doing data analysis in R
is to use [Quarto](https://quarto.org/). Quarto is a way of writing R
(or Python) code alongside text in a way that allows you to create
documents like HTML or Word where the output from R code is directly
inserted into the document.

For example, if you need to make a figure, you can write R code in the
Quarto document so that when you generate a Word document the figure
gets inserted automatically. A main feature of using Quarto is that
because Quarto will always run the code in the document in the same
order and in an empty environment (a new R session) each time you render
the document, the code and results will always be reproducible.

::: callout-tip
If you have heard of [R Markdown](https://rmarkdown.rstudio.com/),
[Quarto](https://quarto.org/) is the next generation version of that.
:::

A Quarto file is a file format (a plain text format like R scripts,
`.csv`, or `.txt` files) with the extension `.qmd` where you write text
with [Markdown](https://pandoc.org/index.html) syntax. Markdown is a
[markup syntax](https://en.wikipedia.org/wiki/Markup_language) and
formatting tool, like HTML, that allows you to write a document in plain
text (e.g. like a `.txt` or `.csv` file). The Markdown text can then be
converted into a vast range of other document types, e.g. HTML, PDF,
Word documents, slides, posters, or websites. In fact, this website is
built from Quarto! Check out Quarto's
[Gallery](https://quarto.org/docs/gallery/) to see a list of things you
can create.

For now, we're going to focus on the main reason that Quarto is used: to
use R code and insert the R output into a document. By using R code in a
document, you can easily switch between data analysis and
document-writing. Which means that:

-   There is less time between exploring a new dataset or analysis and
    sharing your findings with collaborators, because the writing and
    documenting is mixed in with your code for analysis.
-   If you have already created a report and later get new data or find
    out there are problems with the existing data, updating your report
    is as easy as clicking a button to regenerate the results.
-   How you found and present your results is based on the exact
    sequence of steps given in your Quarto document, so showing others
    how the analysis was done is easy because the *how* is explicitly
    shown in the document.
-   Likewise, by reading others' Quarto documents, it is easier to learn
    what was done in their analysis because the logic and sequence is
    shown in the document itself.

{{< text_snippet sticky_up >}}

## Creating a Quarto file

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Briefly explain what the Command Palette is as you use it here. Talk
through this section slowly, especially the YAML header section.
:::

Now, we will create and save a Quarto file. We'll use the Command
Palette by using {{< var keybind.palette >}} and then typing "quarto"
and select the one that says "Create a new Quarto document". You can
also use the menu, by going to
`File -> New File -> Quarto Document ...`, and a dialog box will then
appear. Enter "Reproducible documents" in the title field and your name
in the author field. HTML should be automatically selected as the output
format. There's also the option to use the "visual mode". This mode is
great if you are used to working with Word and you can test it out on
your own later. For this course, we will focus on using the normal mode.

After clicking "Create", the new file will open in RStudio. Before
continuing, let's save this file as `learning.qmd` in the `docs/`
folder.

In the newly saved `docs/learning.qmd` file, you will see some text that
gives a brief overview of how to use the Quarto file. For now, let's
ignore the text. At the top of the file you will see something that
looks a bit like this:

``` {.yaml filename="docs/learning.qmd"}
---
title: "Reproducible documents"
author: "Your Name"
format: html
---
```

This section is called the [*YAML*](https://en.wikipedia.org/wiki/YAML)
*header* and it contains the metadata about the document and the
settings for how Quarto should process it into another document. Most
Markdown documents have this YAML header at the top of the document and
they are always surrounded by `---` on the top and bottom of the
section.

YAML is a data format that has the form of a `key: value` pairing to
store data. The keys in this case are `title`, `author`, and `format`.
The values are those that follow the key (e.g. "Your Name" for
`author`). In the case of Quarto, these `key` data are used to store the
settings that Quarto will use to create the `format` output document.
The keys listed above are some of many settings that Quarto has
available to use.

In the case of this YAML header, the Quarto document will generate an
HTML file because of the `format: html` setting. You can also create a
word document by changing this to `format: docx`.

::: callout-tip
It is possible to create PDF documents, though this requires installing
a LaTeX distributin such as [tinytex](https://yihui.name/tinytex/),
which can sometimes be complicated to install.
:::

So, how do we create a HTML (or Word) document from the this document?
We do that by "rendering" it. At the top of the pane near the "Save"
button, there is a button with the word "Render". To render, you either
click that button or use the shortcut {{< var keybind.render >}}
anywhere in the Quarto document. When you click the "Render" button, a
bunch of processing messages should appear in a new pane beside the
Console, followed by a new window popping up or in the Viewer pane with
the newly created document or.

You've now created a HTML document! Let's try making a Word document.
Change the YAML value in the key `format:` from `html` to `docx`. Then
render the document again with the "Render" button or with the
keybinding {{< var keybind.render >}}. A Word document should open up if
you have a Word processer installed. This is the basic approach to
creating documents from Quarto.

## RStudio layout and usage

You've already gotten a bit familiar with RStudio in the pre-course
tasks, but if you want more details, RStudio has a great [cheat
sheet](https://www.rstudio.com/resources/cheatsheets/#ide) on how to use
RStudio. The items to know right now are the "Console", "Files"/"Help",
and "Source" tabs.

Code is written in the "Source" tab, where it saves the code and text as
a file. You can send selected code to the Console from the opened file
by typing {{< var keybind.run-code >}} (or clicking the "Run" button).
In the "Source" tab (where R scripts and Quarto files are shown), there
is a "Document Outline" button (top right beside the "Run" button) that
shows you the headers or "Sections" (more on that later). To open it you
can either click the button, use the keybinding
{{< var keybind.outline >}}, go through the menu to
`Code -> Show Document Outline`. The Command Palette is a very useful
tool to learn, since you can easily access almost all features and
options inside RStudio through it. Because of this reason, we will be
using it a lot throughout the course. Open it up with
{{< var keybind.palette >}} and then in the pop-up search bar, type out
"document outline". The first item should be the one we want, so hit
`Enter` to activate the Outline.

If you can't remember a specific keybinding in RStudio, check out the
help for it by going to the menu item `Help -> Keyboard Shortcuts Help`.

## Writing R code

Being able to insert R code directly into a document is one of the most
powerful features of Quarto. This frees you from having to switch
between programs when simultaneously writing text and running R code to
derive output that you'd then put into your scientific document.

Running and including R code in Quarto is done using "R code chunks".
You insert these chunks into the document by placing the cursor at the
location where you want the chunk to be, then using the shortcut
{{< var keybind.chunk >}}. Alternatively, you can also use the Command
Palette with {{< var keybind.palette >}} followed by type "new chunk" or
from the menu item `Code -> Insert Chunk`.

Before we insert the code chunk, let's delete all the text in your
document, except for the YAML header (including the dashes surrounding
it). Make sure that the YAML key `format:` is set to `html`. Then, place
your cursor two lines below the YAML header and insert a code chunk with
{{< var keybind.chunk >}}. In the code chunk, type out `2 + 2`. It
should look something like:

```{{r}}
2 + 2
```

You can run R code inside the code chunk using
{{< var keybind.run-code >}} on the line, which will send the code
`2 + 2` to the R Console, with the output appearing directly below the
code chunk in the document. Note that this output is *temporary*.

To see how the output is inserted into the HTML document, let's render
the document using {{< var keybind.render >}} to see what happens. The
output `4` should appear below the code chunk in the HTML document,
something like this:

```{r}
#| echo: fenced
2 + 2
```

## :book: Reading task: Basics of R

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Let them read it over, then briefly go over the content again. We don't
need to do most of this as a code-along, since we will be using them a
lot throughout the course.

Emphasize that, in general, code with `()` means it is a function and
that it does an action. Mention that, like everything, there are some
situations where that isn't completely true but it mostly is.
:::

**Time: \~5 minutes**

Before moving on, let's go over a bit about how R works, and what the "R
session" means. An R session is the way you normally interact with R,
where you would write code in the Console to tell R to do something.
Normally, when you open an R session without an R Project, the session
defaults to assuming you will be working in the `~/Desktop` or `~` (your
Home folder) location. But this location usually isn't where you
actually work and where your R code is. You normally work in the folder
that has your R scripts, Quarto documents, or data files. The assumption
with R Projects on the other hand, is that the R session's working
directory should be where the R Project is, since that is where you have
your R scripts and data files.

In R, everything is an object and every action is a function. A function
is also an object, but an object isn't always a function. To create an
object, also called a variable, we use the `<-` assignment operator. So,
if we want to create an object called `weight_kilos` and assign it the
value `100`, we would write:

```{r using-assignment-operator}
weight_kilos <- 100
weight_kilos
```

The new object now stores the value we assigned it. We can read it like:

> "`weight_kilos` contains the number 100" or "put 100 into the object
> `weight_kilos`"

You can name an object in R almost anything you want, but it's best to
stick to a [style guide](https://style.tidyverse.org/). For example, we
will use `snake_case` to name things.

There are also several main "classes" (or types) of objects in R: lists,
vectors, matrices, and data frames. For now, the only two we will cover
are *vectors* and *data frames*. A vector is a string of values, while a
data frame is multiple vectors put together as columns. Data frames are
a form of data that you'd typically see as a spreadsheet. This type of
data is called "rectangular data" since it has two dimensions: columns
and rows.

So these are vectors, which have different types like character, number,
or factor:

```{r vectors, eval=FALSE}
# Character vector
c("a", "b", "c")
# Logic vector
c(TRUE, FALSE, FALSE)
# Numeric vector
c(1, 5, 6)
# Factor vector (special types of character vectors)
factor(c("low", "high", "medium", "high"))
```

Notice how we use the `#` to write comments or notes. Whatever we write
after the "hash" (`#`) tells R to ignore it and not run it.

This is what a data frame looks like, if we look at the built-in dataset
called `airquality`, which is a data frame object loaded by default when
you start R:

```{r dataframes}
head(airquality)
```

The `c()` function puts values together and `head()` prints the first 6
rows. Both `c()` and `head()` are *functions* since they do an *action*
and they can be recognized by the `()` at their end. Functions take an
*input* (known as *arguments*) and give back an *output*. Each argument
is separated by a comma `,`. Some functions can take unlimited number of
arguments (like `c()`). Others, like `head()` can only take a few
arguments. In the case of `head()`, the first argument requires a data
frame object.

{{< text_snippet sticky_up >}}

## Adding the special "setup" code chunk

We showed a very basic example of how a code chunk works. Let's continue
by doing something slightly more complicated.

One of the major strengths of R, and many other programming languages,
is in its ability for other people to create packages that simplify
doing complex tasks. For example, if you need to use mixed effects
models for your data analysis, you can use the `{lme4}` package. Or if
you want to create figures you can use the `{ggplot2}` package. As you
experienced from the pre-course tasks, installing packages is easy by
using `install.packages()`. Whenever we work with R, we very rarely work
only with the base R functions. We usually use a lot of functions from
many other packages, because that is one of the easiest ways for you to
simplify your work! No need to re-invent the wheel :grin:

One "meta-package" we will use throughout the course is called
`{tidyverse}`. So let's load the package up so we can use the functions
from inside it.

First, go to the top of the `docs/learning.qmd` file and create a new
code chunk two lines below the YAML header with
{{< var keybind.chunk >}}. In the newly created code chunk, type out
`setup` right after the `r`. This area after the `r` is where you write
the code chunk *label*. In this case, we labeled the code chunk with the
name `setup`. Code [chunk
labels](https://yihui.name/knitr/options/#chunk-options) should be named
without `_`, spaces, or `.` and instead should be one word or be
separated by `-`.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Emphasize the warning note below.
:::

::: callout-warning
If you use a space in your chunk label, an error or warning may not
necessarily occur, but there can be unintended side effects that you may
not realize. This may likely cause quite a bit of annoyance and
frustration.

You also can't use duplicate code chunk labels in your document.
:::

::: {.callout-tip appearance="default"}
A nifty thing about using chunk labels is that you can get an overview
of your code chunks using the "Document Outline" with
{{< var keybind.outline >}}, but only if you have this option set up in:
`Tools -> Global Options -> R Markdown -> Show in document outline`.
:::

The name `setup` also has a special meaning in Quarto. When you run
other code chunks in the document in a new (or restarted) session,
Quarto will know to first look for and run the code in the `setup`
chunk. So this is the perfect place to add packages to use or code to
load your data, which we will do in a later session.

The way you load a package and get access to the functions inside is by
using the `library()` function. So let's load the `{tidyverse}` package
by writing `library(tidyverse)` in the code chunk. It should look like
this:

```{{r setup}}
library(tidyverse)
```

Let's run this code chunk by placing the cursor over the code and using
{{< var keybind.run-code >}}. After you run the code, you should see
some text below the `setup` chunk that might look something like this:

```{r}
#| echo: false
#| message: true
#| warning: true
library(tidyverse)
```

As we continue adding code to the `docs/learning.qmd` file, we'll also
add some text to help organize the document and maybe you'll also find
it useful to add notes to yourself as you are coding along. In order for
us to do that, we need to learn a bit about Markdown syntax.

## :book: Reading task: Formatting text with Markdown

**Time: \~8 minutes**

Formatting text in an output document like HTML or Word when using
Markdown is done with the use of "special" characters or syntax. These
special characters control whether the text will be bold, if it will be
a header or a list, and so on. Almost every feature you will need to
write a scientific document is available in Markdown, but it can't do
*everything*. If you can't get Markdown to do what you want, our
suggestion would be to try to fit your writing around Markdown, rather
than force or fight with Markdown to do something it wasn't designed to
do. You might actually find that the simpler Markdown approach is easier
than what you wanted or were thinking of doing, and that you can do
quite a lot with Markdown's capabilities.

::: callout-tip
You can access a quick guide to formatting features of Markdown using
the RStudio menu: `Help -> Cheat Sheets -> R Markdown Cheat Sheet`.
Quarto also has a great guide to the [Basics of
Markdown](https://quarto.org/docs/authoring/markdown-basics.html).
:::

### Headers

Creating headers (like chapters or sections) is done by using one or
more `#` at the beginning of a line followed by some text. Headers
should always be preceded and followed by an empty line:

::::: columns
::: column
``` markdown
# Header 1

Paragraph.

## Header 2

Paragraph.

### Header 3

Paragraph.
```
:::

::: column
# Header 1 {.unnumbered}

Paragraph.

## Header 2 {.unnumbered}

Paragraph.

### Header 3 {.unnumbered}

Paragraph.
:::
:::::

### Lists

Lists are created by adding either `-` or `1.` to the beginning of a
line. An empty line must be at the start and end of the list.

For unnumbered lists, it looks like:

::: {layout-ncol="2"}
#### Markdown {.unnumbered}

``` markdown
- item 1
- item 2
- item 3
```

#### Output {.unnumbered}

-   item 1
-   item 2
-   item 3
:::

And numbered lists look like:

::: {layout-ncol="2"}
#### Markdown {.unnumbered}

``` markdown
1. item 1
2. item 2
3. item 3
```

#### Output {.unnumbered}

1.  item 1
2.  item 2
3.  item 3
:::

### Text formatting

| Markdown        | Output        |
|-----------------|---------------|
| `**bold**`      | **bold**      |
| `*italics*`     | *italics*     |
| `super^script^` | super^script^ |
| `sub~script~`   | sub~script~   |

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Practice using Markdown

**Time: \~10 minutes.**

Get some practice writing Markdown by completing these tasks in the
`docs/learning.qmd` file.

1.  Create one level 2 header (`##`) called "About me" below the `setup`
    code chunk.
2.  Below this new level 2 header, insert a list (either numbered or
    unnumbered) with your name and affiliation(s)/institution(s).
3.  Below the list, write one or two simple sentences about yourself.
    Include one word in **bold** and another in *italics*.
4.  Create another level 2 header (`##`) called "Simple code".
5.  Insert a code chunk below this level 2 header that multiplies
    `3 * 3` and run it.

Use this scaffold to help guide you by replacing the `___` with the
instructions from above:

```` markdown
## ___

-   ___
-   ___

___

## ___


```{{r}}
___
```
````

When you are done, render the document to see how it looks.

{{< text_snippet sticky_up >}}

## Summary

-   Making your research reproducible not only improves the scientific
    quality of your work, but also makes you more efficient, productive,
    and have more confidence in your results.
-   The R session is where you interact with R and where code gets run,
    which occurs in the R Console in RStudio.
-   The working directory in the R Console of the session determines
    where code execution happens, where files are saved, and where data
    is loaded from.
-   Running R code involves sending it to the R Console by using
    {{< var keybind.run-code >}}.
-   There are many types of objects in R, the two important ones for us
    are vectors and data frames.
-   There are many data types in R, like character, numeric, and factor.
-   Everything in R is an object and every action is a function, which
    has a `()` at the end.
-   Use Quarto to create documents that can be turned into a variety of
    file types such as HTML or Word, while also ensuring the analysis is
    (more likely to be) reproducible.
-   Insert R code chucks in Quarto and automatically include the results
    in the final document.
-   Use Markdown syntax like headers (`# Header 1`), text formatting
    (`**bold**`) and lists (`-`) in the Quarto file to format the text
    in the output document.
