# Analytically reproducible documents {#sec-reproducible-documents}

## Learning objectives

This session’s **overall learning outcome** is to:

1.  

**Specific objectives** are to:

1.  Learn what a reproducible document is, how Quarto helps with
    reproducibility, and why it can save you time and effort.
2.  Write and use R code within a document, so that it will
    automatically insert the R output into the final document.
3.  Learn about and use Markdown formatting and syntax for writing
    documents.
4.  Learn about and create different document types like HTML or Word
    from a Quarto document. (TODO, this should be a reading task only)
5.  Become familiar with the very basics of R.
6.  Apply tools to use a consistent "grammar" and "styling" when writing
    R code and making files.

## Why try to be reproducible?

::: callout-note
## Reading task: \~5 minutes

Both reproducibility and replicability are cornerstones for doing
rigorous and sound science. As we've learned, reproducibility in science
is lacking, which this course aims to address. However, being
reproducible isn't just about doing better science. It can also mean
that:

1.  You are much more efficient and productive, as less time is spent
    between coding and transferring your results to a document. No need
    to copy and paste!
2.  You can be confident in your results, since what you report and show
    as figures or tables will be exactly what you get from your
    analysis. Again, no copying and pasting required!

Hopefully by the end of this session, you'll want to start using R
Markdown files for writing your manuscripts and other technical
documents. Believe us, you can save so much time and make your work more
reproducible once you learn how to incorporate text with R code. Plus,
you can create some very aesthetically appealing reports, which are way
easier to produce than if you had done it in Word.

![Have a more reproducible workflow by using Quarto. Artwork by
\@allison_horst.](../images/art-reproducibility.png){width="70%"}
:::

## What is Quarto?

::: callout-note
## Reading task: \~5 minutes

[R Markdown](https://rmarkdown.rstudio.com/) is a file format (a plain
text format like R scripts or `.csv` files) that allows you to be more
reproducible in your analysis and to be more productive in your work. R
Markdown is an extension of [Markdown](https://pandoc.org/index.html)
that integrates R code with written text (as Markdown formatting).

[Quarto](https://quarto.org/) is a next generation version of R Markdown
and chances are, if you've been using a fairly recent version of
RStudio, you are already using it without realizing it. That's because
Quarto uses the same Markdown syntax as R Markdown. The only difference
is that with Quarto, you can create more types of output documents (like
books, websites, slides), you have more options for customization, and
it's easier to do and learn than R Markdown.

So, what is Markdown? It is a [markup
syntax](https://en.wikipedia.org/wiki/Markup_language) and formatting
tool, like HTML, that allows you to write a document in plain text. That
text can then be converted into a vast range of other document types,
e.g. HTML, PDF, Word documents, slides, posters, or websites. In fact,
this website is built from Quarto! The Markdown used in Quarto is based
on [pandoc](https://pandoc.org/index.html) ("pan" means all and "doc"
means document, so "all documents"). Pandoc is a very powerful, popular,
and well-maintained software tool for document conversion. You can use
Quarto to do any number of things. Check out Quarto's
[Gallery](https://quarto.org/docs/gallery/) to see a list of things you
can create. Just a few example document types could be:

-   A website or blog.
-   A PhD thesis.
-   Journal articles.
-   Books (like this course).

For now, we're going to focus on the main reason that Quarto is used: to
incorporate R code and output into a single document. By using R code in
a document, you can have seamless integration between data analysis and
document-writing.

Why would you use this? There are many reasons, with some of them being:

-   There is less time between exploring a new dataset or analysis and
    sharing your findings with collaborators, because the writing and
    documenting is woven in with your code for analysis.
-   If you have already produced a report and later get new data or find
    out there are problems with the existing data, updating your report
    is as easy as clicking a button to regenerate the results.
-   How you found and present your results is based on the exact
    sequence of steps given in your Quarto document, so showing others
    how the analysis was done is easy because the *how* is explicitly
    shown in the document.
-   Likewise, by reading others' Quarto documents, it is easier to learn
    what was done in their analysis because the logic and sequence is
    shown in the document itself.
:::

## Creating a Quarto file

Now, we will create and save a Quarto file. We'll use the Command
Palette by using {{< var keybind.palette >}} and then typing "quarto"
and select the one that says "Create a new Quarto document". You can
also use the menu, by going to
`File -> New File -> Quarto Document ...`, and a dialog box will then
appear. Enter "Reproducible documents" in the title field and your name
in the author field. HTML should be automatically selected as the output
format. There's also the option to use the "visual mode". This mode is
great if you are used to working with Word and you can test it out on
your own later. For this course, we will focus on using the normal mode.

After clicking "Create", the new file will open in RStudio. Before
continuing, let's save this file as `learning.qmd` in the `doc/` folder.

In the newly saved `doc/learning.qmd` file, you will see some text that
gives a brief overview of how to use the Quarto file. For now, let's
ignore the text. At the top of the file you will see something that
looks a bit like this:

``` {.yaml filename="doc/learning.qmd"}
---
title: "Reproducible documents"
author: "Your Name"
format: html
---
```

This section is called the [*YAML*](https://en.wikipedia.org/wiki/YAML)
*header* and it contains the metadata about the document and the
settings for how to process it into another document. Most Markdown
documents have this YAML header at the top of the document and they are
always surrounded by `---` on the top and bottom of the section. YAML is
a data format that has the form of a `key: value` pairing to store data.
The keys in this case are `title`, `author`, and `format`. The values
are those that follow the key (e.g. "Your Name" for `author`). In the
case of Quarto, these `key` data are used to store the settings that
Quarto will use to create the `format` output document. The keys listed
above are some of many settings that Quarto has available to use.

In the case of this YAML header, the Quarto document will generate an
HTML file because of the `format: html` setting. You can also create a
word document by changing this to `format: docx`. You can also create
PDF documents, though this requires installing LaTeX through the R
package [tinytex](https://yihui.name/tinytex/r/), which can sometimes be
complicated to install. We will only cover HTML and Word documents in
this course.

So, how do we create a HTML (or Word) document from the this document?
We do that by "rendering" it. At the top of the pane near the "Save"
button, there is a button with the word "Render" (if it's R Markdown, it
will be called "Knit" with a yarn symbol beside it). To render, you
either click that button or use the shortcut {{< var keybind.render >}}
anywhere in the Quarto document.

When you click the "Render" button, a bunch of processing messages
should appear in a new pane beside the Console, followed by a new window
popping up with the newly created document. Alternatively, the HTML
document may pop up in the "Viewer" pane.

You've now created a HTML document! Let's try making a Word document.
Change the YAML value in the key `format:` from `html` to `docx`. Then
render the document again with the "Render" button or with the
keybinding {{< var keybind.render >}}. A Word document should open up.
This is the basic approach to creating documents from R Markdown or
Quarto. Before continuing, let's add and commit the newly created file
into the Git history with {{< var keybind.git >}}.

## Inserting R code into your document

Being able to insert R code directly into a document is one of the most
powerful features of Quarto. This frees you from having to switch
between programs when simultaneously writing text and running R code to
derive output that you'd then put into your manuscript.

Running and including R code in Quarto is done using "R code chunks".
You insert these chunks into the document by placing the cursor at the
location where you want the chunk to be, then using the shortcut
{{< var keybind.chunk >}}. With the Command Palette, you select the
option from the menu to insert a new code chunk. You can also use the
menu item `Code -> Insert Chunk` to insert a new code chunk.

Before we insert the code chunk, let's delete all the text in your
document, with exception of the YAML header (including the dashes
surrounding it). Make sure that the YAML key `format:` is set to `html`.
Then, place your cursor two lines below the YAML header and insert a
code chunk with {{< var keybind.chunk >}}. In the code chunk, type out
`2 + 2`. It should look something like:

```{r}
#| echo: fenced
2 + 2
```

You can run R code inside the code chunk the same way as you would write
it in an R script. Typing {{< var keybind.run-code >}} on the line will
send the code `2 + 2` to the console, with the output appearing directly
below the code chunk in the document. Note that this output is
*temporary*.

To ensure that the output is inserted into the HTML document, render the
document using {{< var keybind.render >}} and see what happens in the
resulting HTML document. The output `4` should appear below the code
chunk in the HTML document, something like this:

```{r}
2 + 2
```

This is a very simple example of how code chunks work. Things are
usually more complicated than this though. Normally, we have to load R
packages to use for our subsequent code, and this is no different in an
Quarto document. We will set this up together now.

Create a new code chunk with {{< var keybind.chunk >}} and then type
`setup` right after the `r`. This area that you just typed 'setup' in is
for code chunk labels. In this case, we labelled the code chunk with the
name `setup`. Code [chunk
labels](https://yihui.name/knitr/options/#chunk-options) should be named
without `_`, spaces, or `.` and instead should be one word or be
separated by `-`. An error may not necessarily occur if you don't follow
this rule, but there can be unintended side effects that you may not
realize and R will likely not tell you about it, probably causing you
quite a bit of annoyance and frustration. **Note**, you can't use
duplicate code chunk labels in your document.

::: {.callout-tip appearance="default"}
A nifty thing about using chunk labels is that you can get an overview
of your code chunks using the "Document Outline" with
{{< var keybind.outline >}}, but only if you have this option set up in:
`Tools -> Global Options -> R Markdown -> Show in document outline`.
:::

The name `setup` also has a special meaning for Quarto. When you run
other code chunks in the document, Quarto will know to first look for
and run the code in the `setup` chunk. Therefore, this is a good place
to put your `library()` calls or other setup functions. Let's enter some
code to load the packages and the dataset we have been using to the
setup chunk:

```{r setup}
#| echo: fenced
library(tidyverse)
library(NHANES)
nhanes_small <- read_csv(here::here("data/nhanes_small.csv"))
```

Let's insert another code chunk below this one with
{{< var keybind.chunk >}}, and simply put `nhanes_small` in the chunk:

```{r}
#| echo: fenced
nhanes_small
```

Let's run this code as we normally would in a script file, by placing
the cursor over the code and using the shortcut
{{< var keybind.run-code >}}. We can also render the document with
{{< var keybind.render >}} and see what it looks like. When the HTML
document opens, you should see some text below the `setup` chunk that
might look something like this:

```         
── Attaching packages ──────────────────────── tidyverse 1.3.2 ──
✔ ggplot2 3.4.0      ✔ purrr   1.0.0 
✔ tibble  3.1.8      ✔ dplyr   1.0.10
✔ tidyr   1.2.1      ✔ stringr 1.5.0 
✔ readr   2.1.3      ✔ forcats 0.5.2 
── Conflicts ─────────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
```

You probably don't want this text in your generated document, so we will
add a [chunk
option](https://quarto.org/docs/reference/cells/cells-knitr.html) to
remove this message. Chunk options are used to change how code chunks
work. When adding them inside the code chunk, they always need to start
with `#|`. If you want to run the code but not show those messages and
warnings, you can add the options `#| message: false` and
`#| warning: false`:

```{r setup}
#| echo: fenced
#| message: false
#| warning: false
library(tidyverse)
library(NHANES)
nhanes_small <- read_csv(here::here("data/nhanes_small.csv"))
```

If you want to hide the code, messages, warnings, and output, but still
run the code, you can use the option `#| include: false`.

```{r setup}
#| echo: fenced
#| include: false
library(tidyverse)
library(NHANES)
nhanes_small <- read_csv(here::here("data/nhanes_small.csv"))
```

Other common options are:

-   `echo`: To show the code. Default value is `true`. Use `false` to
    hide.
-   `results`: To show the output. Default is `markup`. Use `hide` to
    hide or `asis` as regular text (not inside a code block).
-   `eval`: To evaluate (run) the R code in the chunk. Default value is
    `true`, while `false` does not run the code.

These options all work on the individual code chunk. If you want to set
an option to all the code chunks (e.g. to hide all the code but keep the
output), you can use Quarto's
[`execute`](https://quarto.org/docs/reference/formats/html.html#execution)
options. These options are added to the YAML header and will apply the
settings to everything in the document. We *won't* do this in this
session, but here is what it looks like:

``` {.yaml filename="doc/learning.qmd"}
---
title: "Reproducible documents"
author: "Your Name"
format: html
execute:
  echo: false
  warning: false
  message: false
---
```

## Formatting text with Markdown syntax

::: callout-note
## Reading task: \~8 minutes

Formatting text in Markdown is done using characters that are considered
"special" and act like commands. These special characters indicate what
text is bolded, what is a header, what is a list, and so on. Almost
every feature you will need to write a scientific document is available
in Markdown, although some are missing. If you can't get Markdown to do
what you want, our suggestion would be to try to fit your writing around
Markdown, rather than force or fight with Markdown to do something it
wasn't designed to do. You might actually find that the simpler Markdown
approach is easier than what you wanted or were thinking of doing, and
that you can actually do quite a lot with Markdown's capabilities.

You can access a quick guide to formatting features of Markdown using
the RStudio menu: `Help -> Cheatsheets -> R Markdown Cheat Sheet`.
Quarto also has a great guide to the [Basics of
Markdown](https://quarto.org/docs/authoring/markdown-basics.html).

### Headers

Creating headers (like chapters or sections) is done by using one or
more `#` at the beginning of a line. Headers should always be preceded
and followed by an empty line:

``` markdown
# Header 1

Paragraph.

## Header 2

Paragraph.

### Header 3

Paragraph.
```

### Lists

Lists are created by adding either `-` or `1.` to the beginning of a
line and an empty line must be at the start and end of the list.

For unnumbered lists, it looks like:

``` markdown
- item 1
- item 2
- item 3
```

which gives...

-   item 1
-   item 2
-   item 3

And numbered lists look like:

``` markdown
1. item 1
2. item 2
3. item 3
```

which gives...

1.  item 1
2.  item 2
3.  item 3

### Text formatting

-   `**bold**` gives **bold**.
-   `*italics*` gives *italics*.
-   `super^script^` gives super^script^.
-   `sub~script~` gives sub~script~.

### Inline R code

Quarto also allows you to include output in-text. For instance, if you
wanted to add the mean of some values to the text, it would look like
this:

> The mean of BMI is
> `` `r knitr::inline_expr("round(mean(nhanes_small$bmi, na.rm = TRUE), 2)")` ``.

which gives...

> The mean of BMI is `r round(mean(nhanes_small$bmi, na.rm = TRUE), 2)`.

But note that using inline R code can *only* insert a *single* number or
character value, and nothing more.
:::

::: {.callout-tip appearance="default"}
> For more details about other Markdown "syntax", check out
> @sec-extra-markdown as well as the R Markdown cheatsheet
> (`Tools -> Cheatsheets`, which works for many Quarto features too) and
> Quarto's [Markdown
> Basics](https://quarto.org/docs/authoring/markdown-basics.html) page.
> Continue to the exercise below.
:::

## Exercise: Practice using Markdown for writing text

> Time: \~5 minutes.

Get some practice writing Markdown by completing these tasks in the
`doc/learning.qmd` file.

-   Right under the YAML header, insert a list (either numbered or
    unnumbered) with your name and affiliation(s)/institution(s).
-   Create three level 1 headers (`#`), called "Intro", "Methods and
    Results", and "Discussion".
-   Create a level 2 header (`##`) under "Methods and Results" called
    "Analysis".
-   Write one random short sentence under each header. Bold (`**word**`)
    one word in each and italicize (`*word*`) another.
-   Insert a code chunk to make a simple calculation (e.g. `2 + 2`).
-   Finally, add and commit these changes into the Git history with
    {{< var keybind.git >}}.

## RStudio layout and usage

Open up the `R/learning.R` file now, which you will use to type in code
for the code-along parts. You've already gotten a bit familiar with
RStudio in the pre-course tasks, but if you want more details, RStudio
has a great
[cheatsheet](https://www.rstudio.com/resources/cheatsheets/#ide) on how
to use RStudio. The items to know right now are the "Console",
"Files"/"Help", and "Source" tabs.

Code is written in the "Source" tab, where it saves the code and text as
a file. You can send selected code to the Console from the opened file
by typing {{< var keybind.run-code >}} (or clicking the "Run" button).
In the "Source" tab (where R scripts and Quarto files are shown), there
is a "Document Outline" button (top right beside the "Run" button) that
shows you the headers or "Sections" (more on that later). To open it you
can either click the button, use the keybinding
{{< var keybind.outline >}}, go through the menu to
`Code -> Show Document Outline`. The Command Palette is a very useful
tool to learn, since you can easily access almost all features and
options inside RStudio through it. Because of this reason, we will be
using it a lot throughout the course. Open it up with
{{< var keybind.palette >}} and then in the pop-up search bar, type out
"document outline". The first item should be the one we want, so hit
`Enter` to activate the Outline.

If you can't remember a specific keybinding in RStudio, check out the
help for it by going to the menu item `Help -> Keyboard Shortcuts Help`.

## Basics of using R

One useful thing to do to make your R script more readable and
understandable is to use "Sections". They're like "headers" in Word and
they split up an R script into sections, which then show up in the
"Document Outline". We can insert a section using
{{< var keybind.code-section >}}. You can also insert the sections
through the menu `Code -> Insert Section`.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Let them read it over, then briefly go over the content again. We don't
need to do most of this as a code-along, since we will be using them a
lot over the later sessions. However, do a code-along showing how to
assign data to objects, the difference between unassigned (not saved)
and assigned (saved; this will be helpful in the wrangling section and
piping without assigning), and how to send code to the Console.

Emphasize that, in general, code with `()` means it is a function and
that it does an action. Mention that, like everything, there are some
situations where that isn't completely true but it mostly is.
:::

::: callout-note
## Reading task: \~5 minutes

In R, everything is an object and every action is a function. A function
is an object, but an object isn't always a function. To create an
object, also called a variable, we use the `<-` assignment operator:

```{r using-assignment-operator}
weight_kilos <- 100
weight_kilos
```

The new object now stores the value we assigned it. We can read it like:

> "`weight_kilos` contains the number 100" or "put 100 into the object
> `weight_kilos`"

You can name an object in R almost anything you want, but it's best to
stick to a [style guide](https://style.tidyverse.org/). For instance, we
will use `snake_case` to name things.

There are also several main "classes" (or types) of objects in R: lists,
vectors, matrices, and data frames. For now, the only two we will cover
are *vectors* and *data frames*. A vector is a string of values, while a
data frame is multiple vectors put together as columns. Data frames are
a form of data that you'd typically see as a spreadsheet. This type of
data is called "rectangular data" since it has two dimensions: columns
and rows.

So these are vectors, which have different types like character, number,
or factor:

```{r vectors, eval=FALSE}
# Character vector
c("a", "b", "c")
# Logic vector
c(TRUE, FALSE, FALSE)
# Numeric vector
c(1, 5, 6)
# Factor vector
factor(c("low", "high", "medium", "high"))
```

Notice how we use the `#` to write comments or notes. Whatever we write
after the "hash" (`#`) tells R to ignore it and not run it.

This is what a data frame looks like:

```{r dataframes}
head(airquality)
```

The `c()` function puts values together and `head()` prints the first 6
rows. Both `c()` and `head()` are *functions* since they do an *action*
and they can be recognized by the `()` at their end. Functions take an
*input* (known as *arguments*) and give back an *output*. Each argument
is separated by a comma `,`. Some functions can take unlimited arguments
(like `c()`). Others, like `head()` can only take a few arguments. In
the case of `head()`, the first argument is reserved for the name of the
data frame.
:::

## Using auto-completion in RStudio

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Really emphasize how use auto-completion is.
:::

To more quickly type out objects in R, use "tab-completion" to finish an
object name for you. Normally RStudio will start auto-completing for you
as you type code, but you can manually trigger auto-completion with
`Tab`. As you type out an object name, hit the `Tab` key to see a list
of objects available. RStudio will not only list the objects, but also
shows the possible options and potential help associated with the
object.

Try it out. In the RStudio Console, start typing:

```{r}
#| filename: "Console"
#| eval: false
col
```

Then hit tab. You should see a list of functions to use. Hit tab again
to finish with `colnames()`. This simple tool can save so much time and
can prevent spelling mistakes.

If we want to get more information from data frames, we can use other
functions like:

```{r examining-dataframes}
#| filename: "Console"
# Column names
colnames(airquality)

# Structure
str(airquality)

# Summary statistics
summary(airquality)
```

## Summary

-   Keep R scripts simple, focused, and short.

-   Use tab auto-completion when writing code.

-   Use `?` to get help on an R object.

-   Making your research reproducible not only improves the scientific
    quality of your work, but also makes you more efficient, productive,
    and have more confidence in your results.

-   Use Quarto to construct files that can easily be turned into a
    variety of file types such as HTML or Word.

-   Insert R code chucks in Quarto and automatically include the results
    in the final document.

-   Use headers (`# Header 1`), text formatting (`**bold**`) and lists
    (`-`) in the Quarto file.
