# Split, apply, and combine {#sec-split-apply-combine}

```{r setup}
#| include: false
library(tidyverse)
post_meal_data <- read_csv2(here::here("data/post-meal-insulin.csv"))
tidied_post_meal_data <- post_meal_data |>
  select(id, Age, BMI, Group, auc_pg, auc_ins, glykemi) |>
  rename(
    auc_glucose = auc_pg,
    auc_insulin = auc_ins
  ) |>
  filter(glykemi == 0) |>
  select(-glykemi) |>
  mutate(
    insulin_glucose_ratio = auc_insulin / auc_glucose,
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  )
```

The split-apply-combine method is a powerful technique for performing
data analysis. It involves splitting a dataset into groups based on some
categorical variable(s), performing some analysis to each group, and
then combining the results back together into a single dataset. In this
session, we'll explore how to use this technique in R using the
`group_by()` and `summarise()` functions from the `{dplyr}` package.

## Learning objectives

{{< include /includes/objectives/_split-apply-combine.qmd >}}

## :book: Reading task: Split-apply-combine to summarize data

**Time: \~5 minutes**

Summarising or applying simple (or complex) statistics to data is a key
component of any analysis. Simple summaries or statistics can be done
either on all of the data or on groups of it. There are many data
analysis tasks that can be approached using the
[split-apply-combine](https://www.jstatsoft.org/article/view/v040i01)
method, which involves splitting the data into groups, applying some
analysis to each group, and then combining the results together.

As we briefly showed in @tbl-wrangling-verbs, `{dplyr}` has some
wrangling verbs that include one to `summarise()` the data. If you want
to do a split-apply-combine analysis to, for example, find the max
height by a group, you would use the functions `group_by()` and then
`summarise()`. Using `group_by()` splits the data up and `summarise()`
applies an analysis, then immediately combines it back together.

The first position argument to `group_by()` is, as usual, the data. The
next arguments are the columns that contain the values you want to group
by. These columns must contain **categorical** data (e.g. education).
`group_by()` tells R to compute the next operations on the data within
each grouping, rather than on all the data.

As with the other functions, `summarise()` takes the data as the first
position argument. The next arguments work similar to the arguments in
`mutate()` with one difference: the output must create a single value
(e.g. a mean). Like `mutate()`, you can add multiple "summaries" by
adding new columns separated by `,`. You would use `summarise()` to
derive basic descriptive statistics of a certain variable, including
`min()`, `max()`, `mean()`, `median()`, or `sd()` (standard deviation).

The `group_by()` function doesn't do anything by itself and only is used
in combination with a `summarise()`, `mutate()`, `arrange()`, or other
function. However, the `summarise()` function can be used on its own.

{{< text_snippet hats_up >}}

## Calculating summary statistics

Let's try using `summarise()` without using `group_by()`. Go to the
bottom of your `docs/learning.qmd` Quarto file and create a new header
called `## Calculating summary statistics`. Then, create a new code
chunk with {{< var keybind.chunk >}}. Let's start with calculating the
maximum value of the BMI variable.

```{r calculate-max-bmi}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  summarise(max_bmi = max(BMI))
```

We should get the max value of BMI in our data.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Verbally explain this next paragraph about `na.rm`, but you don't need
to code it.
:::

Sometimes, if there are missing values in your data, you might get `NA`
as a result, which means "missing". In R, `NA` values "propagate",
meaning that if there is one value missing, then the max or mean value
will also be missing. When this happens, we need to tell `max()` to
exclude any `NA` values from the calculation by using the argument
`na.rm = TRUE`.

To calculate another summary statistic, you would add another summary
column using `,`:

```{r calculate-min-max-bmi}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  summarise(
    max_bmi = max(BMI),
    min_bmi = min(BMI)
  )
```

This is the basics of using `summarise()` to calculate summary
statistics. The real power of this function comes when you combine it
with `group_by()`.

## Summary statistics by a group

While the `summarise()` function is useful enough on its own, it really
shines when combined with `group_by()`. Create a new header called
`## Summarise with group by` and below that add a new code chunk with
{{< var keybind.chunk >}}. Let's calculate the mean age and BMI between
the two groups.

```{r calculate-mean-age-bmi-by-group}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    mean_age = mean(Age),
    mean_bmi = mean(BMI)
  )
```

::: callout-note
You'll likely get a message informing you that it is
`regrouping output`. This is a notification and can be ignored. If you
don't want the message displayed, you can add
`options(dplyr.summarise.inform = FALSE)` to the top of your script and
run it.
:::

Run this code chunk with {{< var keybind.run-code >}} to see the output.
Cool! Since we don't need the dataset grouped anymore, it's good
practice to end the grouping with `ungroup()`.

```{r ungroup-data}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    mean_age = mean(Age),
    mean_bmi = mean(BMI), 
    .groups = "drop"
  ) 
```

Run this code chunk with {{< var keybind.run-code >}} to see the output.
It isn't different from before except it is no longer grouped. Before
continuing, let's run `{styler}` using {{< var keybind.styler >}},
render the document with {{< var keybind.render >}}, then add and commit
changes to the Git history using {{< var keybind.git >}} before finally
pushing to GitHub.

## Creating tables of results

A useful combination of doing a split-apply-combine analysis for
reporting or showing it as a table for easy viewing and sharing with
collaborators. The `knitr` package has a function called `kable()` that
can be used to create tables from R objects. So let's test out making
one!

We'll work from the code we just created from above, but we'll add
`knitr::kable()` to the end of the pipe.

```{r create-knitr-table}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    mean_age = mean(Age),
    mean_bmi = mean(BMI),
    .groups = "drop"
  ) |>
  knitr::kable()
```

Then run the code with {{< var keybind.run-code >}} to see what it looks
like. In RStudio, the output doesn't look too good. But let's render it
with {{< var keybind.render >}} and view the output in the HTML
document. It looks much better! But could use some improvements. For
instance, the columns names could be more "human friendly", the table
could use a caption, and the numbers could be rounded.

Let's start with the rounding. We can use the `round()` function to
round the values to a certain number of digits. We'll add this to the
individual new `summarise()` columns we are making. Let's round to 1
digit for now.

```{r round-values-for-table}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    mean_age = round(mean(Age), 1),
    mean_bmi = round(mean(BMI), 1),
    .groups = "drop"
  ) |>
  knitr::kable()
```

Run this code chunk with {{< var keybind.run-code >}} to see the output.
Next, let's fix the column names. In both `mutate()` and `summarise()`,
if you want to create columns that have specific characters in the name
that R normally doesn't allow or doesn't like, you can `""` around the
name. So in this case, we can put `""` around the column names to make
them more human readable.

```{r names-in-quotes-for-table}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    "Mean Age (yrs)" = round(mean(Age), 1),
    "Mean BMI (kg/m^2^)" = round(mean(BMI), 1),
    .groups = "drop"
  ) |>
  ungroup() |>
  knitr::kable()
```

Run this code chunk with {{< var keybind.run-code >}} to see the output.
To add the caption, there are two ways to do it: as an argument to
`kable()` or using the Quarto code chunk options `#| tbl-cap:` that
works similar to the one for figures. We can also reference the table
too if we include a label with `#| label: tbl-`.

```` {.markdown filename="docs/learning.qmd"}
```{{r}}
#| label: tbl-mean-age-bmi
#| tbl-cap: "Mean values of Age and BMI for each group."
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    "Mean Age (yrs)" = round(mean(Age), 1),
    "Mean BMI (kg/m^2^)" = round(mean(BMI), 1),
    .groups = "drop"
  ) |>
  knitr::kable()
```
````

```{r}
#| label: tbl-mean-age-bmi
#| tbl-cap: "Mean values of Age and BMI for each group."
#| echo: false
tidied_post_meal_data |>
  group_by(Group) |>
  summarise(
    "Mean Age (yrs)" = round(mean(Age), 1),
    "Mean BMI (kg/m^2^)" = round(mean(BMI), 1),
    .groups = "drop"
  ) |>
  knitr::kable()
```

Then, when we use `@tbl-mean-age-bmi` in the text, it will reference the
table like this: @tbl-mean-age-bmi, cool eh! Let's run `{styler}` using
{{< var keybind.styler >}} and then render the document with
{{< var keybind.render >}} to see what it looks like!

Before continuing, add and commit these changes into the Git history
using {{< var keybind.git >}}, followed by pushing to GitHub.

## :technologist: Exercise: Creating a table using R code

**Time: \~15 minutes.**

Let's create a table with more columns and rows, to practice using
split-apply-combine with `knitr::kable()`. In the `docs/learning.qmd`
file, create a new header called `## Bigger table`. Then copy the code
template below and paste it below the new header. Then, using this
template, complete each of the items below.

```` {.markdown filename="docs/learning.qmd"}
```{{r}}
#| label: tbl-___
#| tbl-cap: "___"
post_meal_data |>
  # Task 2.
  ___(BMI < 30) |>
  # Task 3.
  ___(
    ___ = if_else(Age < 40, "Young", "Old"),
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  ) |>
  # Task 4.
  ___(___, ____) |>
  summarise(
    "___" = ___(___(auc_cp), ___),
    "___" = ___(___(auc_pg), ___),
    "___" = ___(___(auc_ins), ___)
  ) |>
  # Task 5.
  ___(
    "___" = age_group,
    "___" = Group
  ) |>
  # Task 6.
  ___() |>
  ___::___()
```

See @tbl-___, very nice table! :D
````

1.  In the code chunk options, set the `#| label:` to be
    `tbl-summary-table` and the `#| fig-cap:` to be
    `"Median values of AUC measurements by age group and family history group."`.
    Or something similar, it doesn't have to be exact.

2.  Use `filter()` to only include rows where `BMI` is less than 30.

3.  Use `mutate()` to create a new column called `age_group` using
    `if_else()`. Since you already created this code as well as
    modifying the `Group`, we've provided it in the template for you, so
    you don't need to add it again.

4.  Calculate the `median()` values of `auc_cp`, `auc_pg`, and `auc_ins`
    by `Group` and `age_group` using `group_by()` and `summarise()`. Use
    the `round()` function to round the values to 1 digit. Remember, the
    `median()` function is used first, *inside* the `round()` function
    (meaning, the `round()` function is on the outside). Name the new
    columns `"AUC c-Peptide"`, `"AUC Glucose"`, and `"AUC Insulin"`
    (note the `""` around the names).

5.  `rename()` the columns to be more human readable. Rename `age_group`
    to `"Age group"` and `Group` to `"Family history"`. Remember,
    renaming follows the `new = old` format.

6.  Next, `ungroup()` the data before sending it to `knitr::kable()` to
    create the table.

7.  In the text below the code chunk, reference the table with
    `@tbl-summary-table`.

8.  Run `{styler}` on the document with {{< var keybind.styler >}}.

9.  Render the document to HTML with {{< var keybind.render >}} to see
    what the table looks like.

10. End the exercise by adding and committing to the Git history with
    {{< var keybind.git >}}, then push to GitHub.

```{r solution-create-pretty-table}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# Code might look like this:
#| label: tbl-summary-table
#| tbl-cap: "Median values of AUC measurements by age group and family history group."
post_meal_data |>
  # Task 2.
  filter(BMI < 30) |>
  # Task 3.
  mutate(
    age_group = if_else(Age < 40, "Young", "Old"),
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  ) |>
  # Task 4.
  group_by(Group, age_group) |>
  summarise(
    "AUC c-Peptide" = round(median(auc_cp), 1),
    "AUC Glucose" = round(median(auc_pg), 1),
    "AUC Insulin" = round(median(auc_ins), 1)
  ) |>
  # Task 5.
  rename(
    "Age group" = age_group,
    "Family history" = Group
  ) |>
  # Task 6.
  ungroup() |>
  knitr::kable()

# See @tbl-summary-table, very nice table! :D
```

{{< text_snippet hats_up >}}

## Summary

-   Use the "split-apply-combine" technique with the functions
    `summarise()` and `group_by()` from the `{dplyr}` package to
    calculate summaries of your data.
-   Make tables by using `knitr::kable()`
-   Use the `#| label: fig-LABELNAME` and `#| fig-cap: "CAPTION"`
    options to label and caption tables, and reference the table in the
    text with `@tbl-LABELNAME`.

## Survey

Please complete the survey for this session:

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
