# Data management and wrangling {#sec-wrangling}

```{r setup}
#| include: false
library(tidyverse)
post_meal_data <- read_csv2(here::here("data/post-meal-insulin.csv"))
```

<!-- TODO: Add an introduction here -->

## Learning objectives

This session's **overall learning outcome** is to:

1.  Describe some basic, common "verbs" for wrangling data, such as
    selecting and renaming columns, filtering rows, and "mutating"
    columns by modifying or creating them. Then use functions within the
    `{tidyverse}` to perform these actions.

**Specific objectives** are to:

1.  Select specific columns in a dataset using the `select()` function.
2.  Rename columns in a dataset using the `rename()` function.
3.  Filter rows in a dataset using the `filter()` function.
4.  Modify or add columns in a dataset using the `mutate()` function.
5.  Use the pipe operator `|>` to chain actions together.

## :book: Reading task: Data wrangling verbs

**Time: \~5 minutes**

A large number of tasks in data wrangling can be broken down and
expressed with only a few simple "verbs" (actions). Wrangling here
refers to maneuvering, managing, controlling, and turning your data
around to clean it up, better understand it, and prepare it for later
analyses. The table below lists some common "verbs" from the `{dplyr}`
package that gets loaded with `{tidyverse}`:

| Task | Function | Example |
|:-----------------------|:-----------------------|:-----------------------|
| Select columns | `select()` | Remove data entry columns such as the person's name who entered the data. |
| Subset/filter rows | `filter()` | Keeping rows with glucose values above 4. |
| Rename columns | `rename()` | Changing a column name from 'Q1' to 'participant_name'. |
| Transform or modify columns | `mutate()` | Multiplying or taking the log of a column's values. |
| Sort rows | `arrange()` | Show rows with the smallest value at the top. We won't cover this function in this course, but we do have an extra section on it in @sec-extra-arrange. |
| Calculate summaries | `summarise()` | Calculating the maximum, median, and minimum age. We'll cover this in @sec-split-apply-combine. |
| Run code by a group | `group_by()` with `summarise()` | Calculate means of age by males and females. We'll cover this in @sec-split-apply-combine. |

: List of common data wrangling tasks, the function used, and an example
of when it would be used.

You might notice that we have used the word "variable" in the previous
sessions and here we are starting to use "column". While often these two
words mean the same, they don't always. A column is a list of values of
the same type structured in a data frame. A variable is a list of values
that have specific and consistent characteristics that represent
something in the real world. For instance, in R, it is entirely possible
to have all the columns in a data frame stacked on top of each other in
a very long format, including stacking numbers and text. A variable
however, would be something like a person's weight or age. It is
something that is important to us humans, that means something specific.
When we do data wrangling, we often end up working with or creating
columns, but often the goal to get our data into a format where the
columns are also the variables.

Taking an example from our data, the `Insulin10` column would be more of
a column than a variable because it is insulin measured at 10 minutes,
which is very specific and isn't really meaningful to us humans as a
concept. If we split it up into two new columns, `Insulin` and `Time`,
then both `Insulin` and `Time` would be variables.

{{< text_snippet sticky_up >}}

## Selecting specific columns in a dataset

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Take your time explaining the arguments and positions of things in the
function.
:::

Selecting columns of a dataset is a very common data wrangling task. The
function for this task is appropriately called `select()`. You would use
`select()` to select one or more columns in a dataset that you want to
have a closer look at or to save as a new data frame to work with. It
may be that you wish to explore the clinical characteristics of your
study sample, so you may select some basic demographic variables (e.g.,
`Age`) and clinical variables (e.g., `Weight` and `Height`) to perform
these analyses.

For the input arguments, `select()` takes the data object as the first
argument, which is the first input position right after the opening
bracket `(`, and then takes the names of the columns you want to select
after this first argument position. The argument after the data argument
is shown as `...`, which indicates that you can add as many columns as
you want, separated by a `,`. Let's try it out on a few columns of our
data.

First, go to the bottom of your Quarto file and create a new header
called `## Selecting data`. Below that header, create a new code chunk
with {{< var keybind.chunk >}}. Then we can start selecting some
columns. Let's select one column first. When you write the name of the
column, you don't need to use quotes around it but it will still work if
you do.

```{r using-select-one-column}
#| filename: "docs/learning.R"
select(post_meal_data, Age)
```

When you run this code with {{< var keybind.run >}}, you will see that
only the `Age` column is shown in the output.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Open up and show the Environment pane to show that no new object has
been created and explain the paragraph below. Reinforce that if you
don't assign the output to an object, it means you can't do anything
else with the output since nothing was created.
:::

If we look at our environment pane in RStudio, you will notice that
running these functions doesn't create any new object in our
environment. That's because we haven't assigned the output to a new
object with `<-`. When we run these actions, we are only telling R to
send the output to our screen for us to look at. Shortly we will assign
a new object with some selected columns to a new data object.

Let's select a few more columns. Create a new line in the same code
chunk and add the following code:

```{r using-select-multiple-columns}
select(post_meal_data, Age, BMI, Weight)
```

This time we have a lot more columns output. What if we want to exclude
a column? We can use the minus sign `-` to exclude a column. Let's
exclude the `Group` column.

```{r using-select-exclude-column}
select(post_meal_data, -Group)
```

Sometimes you might have many columns with that all follow a specific
pattern, like our `Insulin` or `PG` columns. There are some "select"
helper functions to choose these columns, the commonly used ones are:

-   `starts_with()`: Select columns that begin with a pattern.
-   `ends_with()`: Select columns that end with a pattern.
-   `contains()`: Select columns that contain a pattern.

For more information on using the pattern functions such as
`starts_with()`, check out `?select_helpers` (choose the "Select
helpers" option in the menu that pops up).

Let's try these out on our data. Create a new header called
`## Select helpers` and then create a new code chunk below that header
with {{< var keybind.chunk >}}. We'll start by selecting all columns
that start with "PG" (post-meal glucose).

```{r using-select-helpers-starts-with}
#| filename: "docs/learning.qmd"
select(NHANES, starts_with("PG"))
```

See how it outputs all the columns that start with "PG"? Now let's try
selecting all columns that end with "DXA":

```{r using-select-helpers-ends-with}
select(NHANES, ends_with("DXA"))
```

Just like the previous example, it outputs all the columns that end with
"DXA". Lastly, let's try selecting all columns that contain the number
"0" somewhere in the column name:

```{r using-select-helpers-contains}
# All columns containing the number "0"
select(NHANES, contains("0"))
```

Now that we've tested how to use `select()`, let's simplify our dataset
a bit. While our data doesn't have too many rows, there are a lot of
columns to deal with, at least for this course. Let's keep only a few
columns for now, so it is easier to look at and explore. Create a new
header called `## Simplifying data` and then create a new code chunk
below that header with {{< var keybind.chunk >}}. Then write out the
code below, so that we can make a new, smaller data object.

```{r create-smaller-data}
#| filename: "docs/learning.qmd"
small_post_meal_data <- select(post_meal_data, id, Age, BMI, Group, auc_pg, auc_ins, glykemi)

# View the new data frame
small_post_meal_data
```

We will now work with this smaller data object for a while. Let's make
sure to style our document with {{< var keybind.styler >}}, then we will
render the document with {{< var keybind.render >}} to see the output.
Once done, we'll add and commit the changes to the Git history with
{{< var keybind.git >}}. Finally, push the changes to your GitHub
repository.

## :book: Reading task: Chaining actions with the pipe

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Briefly go over this again, especially the examples given.
:::

**Time: \~6 minutes**

A key component of tidy data and tidy code is making use of the "pipe"
`|>` operator and that comes built-in with R. This base R pipe was
inspired from the `{magrittr}` `%>%` pipe. You may notice this style of
piping in documentation online. Both pipes work exactly the same, but
since the `|>` comes built-in with R, we will use that one for this
course.

You would use the pipe `|>` operator when you are writing a piece of
code with multiple operations or steps in sequence. One advantage of the
pipe is that it makes your code a bit more readable and also a bit
easier to modify to insert additional steps in your data analysis
"pipeline".

The pipe takes the output from one function and inserts it as input to
another function, just like a plumbing pipe would do for water. This
allows you to easily chain functions together into "sentences".

Let's use an example based on English words for some action. This is the
English sentence:

> We need some eggs. Drive to the grocery store and buy some eggs before
> coming home from work.

There are basically two actions here ("drive" and "buy") with four
inputs ("work", "grocery store", "eggs", "home"), that are all based on
the previous action. Since an action in R is a function, the functions
would be `drive()` and `buy()`. In regular R (without the "pipe"
operator), we would have to nest functions (reading from the inside to
the outside) together to chain them:

``` r
drive(buy(drive(at_work, "grocery store"), "eggs"), "home")
```

This is a bit difficult to read. Another way to make it more readable
would be to create temporary objects for each step:

``` r
at_grocery_store <- drive(at_work, "grocery store")
got_eggs <- buy(at_grocery_store, "eggs")
at_home <- drive(got_eggs, "home")
```

This still isn't very "readable", as we are having to re-name each
intermediate object with reference to the object before it. The pipe
`|>` operator can really simplify this:

``` r
at_work |> 
    drive("grocery store") |> 
    buy("eggs") |> 
    drive("home")
```

Do you find this more readable and understandable? It reads how it would
be done, in the order of the steps taken.

The idea of piping is to read the functions from left to right. This can
help clarify and break down complex data processing workflows, and is
the basis for the `{tidyverse}` and many other packages.

The pipe `|>` takes the output from the object or function from the left
of the operator and puts it into the function on the right of the
operator. All input goes into the first position argument of the
function. Within the `{tidyverse}` packages, all functions take a data
frame (or vector) as the first argument in order to work with the pipe.

{{< text_snippet sticky_up >}}

## Renaming specific columns

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Take your time to explain the pipe operator and how it works.
:::

Renaming columns is another very common data wrangling task. The
function for this task is also pretty clear, called `rename()`.
Depending on how your data was collected, it may have column names that
aren't very descriptive. So you'll probably want to rename them to
something more explanatory, which is particularly important if you're
sharing your work with others or in an environment where multiple people
are working on the same data.

Like `select()`, `rename()` takes the dataset as the first argument and
then takes as many renaming arguments as you want (because the second
argument position is `...`). When renaming, it takes the form of
`newname = oldname`. Almost all functions in the `{tidyverse}` take this
form, where the first argument is the data object and the rest are the
arguments that the function will use. The purpose of this design is to
make the functions easy to use with the pipe, because the pipe
automatically takes the data object and puts it into the first position
of the function.

Let's rename a few of the columns, while also starting to use the pipe
operator. Create a new header called `## Renaming columns` and then
create a new code chunk below that header with
{{< var keybind.chunk >}}. Then write out the code below, so that we
start using the pipe operator to rename some columns. The keyboard
shortcut for the pipe is {{< var keybind.pipe >}}. Let's rename `auc_pg`
to be a bit clearer as `auc_glucose`.

```{r rename-columns}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  rename(auc_glucose = auc_pg)
```

The pipe automatically takes `post_meal_data` and puts it into the first
position, so we don't need to specify `post_meal_data` inside the
`rename()` function. Run this code with {{< var keybind.run >}} and you
will see the output with the new column name. We also have a column
`auc_ins` that we can rename to `auc_insulin`. Add this to the code
chunk below the previous code.

```{r rename-columns-insulin}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  rename(
    auc_glucose = auc_pg,
    auc_insulin = auc_ins
  )
```

Run this code again with {{< var keybind.run >}} to see the output.
Before moving on, let's style our document with
{{< var keybind.styler >}}, render the document with
{{< var keybind.render >}}, add and commit the changes to the Git
history with {{< var keybind.git >}}, then push to GitHub.

## :technologist: Exercise: Practice selecting and renaming

**Time: \~10 minutes.**

Complete the following tasks while in the `docs/learning.qmd` document,
running each code after completing the task. *Remember that you aren't
actually changing the data each time, since you won't assign (`<-`) to a
new object.* Use the code below as a template to complete the tasks.

``` {.r filename="docs/learning.qmd"}
# Task 1.
post_meal_data |>
  ___(___(___))

# Task 2.
post_meal_data |>
  ___(___(___), -___(___)) |>
  ___(
    ___ = ___,
    ___ = ___
  )

# Task 3.
basic_info <- select(post_meal_data, id, BMI, Length, Weight, Age)
rename(basic_info, Height = Length)
```

Create a new header called `## Select and rename tasks` at the bottom of
the Quarto document. Then create a new code chunk below that header with
{{< var keybind.chunk >}} and copy and paste the template above into
this code chunk. Complete the tasks below by filling in the blanks in
the template:

1.  `select()` only columns that `contain()` the text `"OGTT"`. Run that
    code to get the output. How many columns are there?

2.  `select()` columns that `contain()` the characters `".."` but
    exclude `-` columns that `contain()` the characters `"..."`. Then,
    rename those columns so that instead of `..`, they have `.minus5`.
    Recall that renaming is in the form `new = old`.

3.  Read through (in your head) the code below `# Task 3.` above. How
    intuitive is it to read? Now, re-write this code to use the pipe
    `|>` so that you don't need to create the temporary `basic_info`
    object, then re-read the revised version. Which do you feel is
    easier to "read"?

After you finish these tasks, run `{styler}` on the `docs/learning.qmd`
file with {{< var keybind.styler >}}, then render the document with
{{< var keybind.render >}}. Lastly, add and commit these changes to the
Git history with {{< var keybind.git >}} before pushing to GitHub.

```{r solution-practice-select-rename}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# Task 1.
post_meal_data |>
  select(contains("OGTT"))

# Task 2.
post_meal_data |>
  select(contains(".."), -contains("...")) |>
  rename(
    P.Glucose.minus5.OGTT = P.Glucose..5.OGTT,
    Insulin.minus5.OGTT.X = Insulin..5.OGTT.X
  )

# Task 3.
post_meal_data |>
  select(id, BMI, Length, Weight, Age) |>
  rename(Height = Length)
```

{{< text_snippet sticky_up >}}

## Filtering data by row

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Emphasize the importance of being careful with logic conditions, that we
humans are really bad at it.
:::

The two functions to "subset" are `select()`'ing on the column level,
and `filter()`'ing on the row level. You'll use filtering to do tasks
like removing outliers or incorrectly entered data, dropping missing
values, or keeping only some rows based on a condition, like only
participants in a certain group.

The `filter()` function requires a logic condition (`TRUE` or `FALSE`).
As with the other functions, the first argument is the dataset and all
others are the logical conditions that will apply to the row filtering.
When the logical conditions equal `TRUE`, it means that those rows will
be **kept** and those that are `FALSE` will be *dropped*.

::: callout-warning
Since `filter()` uses logical conditions, you need to be really careful
when writing the logic. As you probably know, humans are really *really*
bad at logic. If your logical condition starts getting complex, double
and triple check that you know for certain that your logic code is doing
what you think it is doing. It's very easy to make mistakes at this
stage, even for advanced R users.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Show this table on the screen and briefly explain the first few to them.
Mention that they can use this table as a reference for the rest of the
course.
:::

Below is a table showing most of the logic comparisons to use in R.
@tbl-logic-operators can be used as a reference for logical conditions
in R.

| Operator | Description                                  |
|:---------|:---------------------------------------------|
| \<       | less than                                    |
| \<=      | less than or equal to                        |
| \>       | greater than                                 |
| \>=      | greater than or equal to                     |
| ==       | equal to                                     |
| !=       | not equal to                                 |
| !x       | not x (if x is true, `!` will make it false) |
| x \| y   | x OR y                                       |
| x & y    | x AND y                                      |

The simplest kind of logic condition is to test for "equality". In R,
"equal to" is represented by `==`. For example, if we wanted to keep
only people who are have (hyper-)glycemia in the dataset, we would use
the logic condition like this. First, go to the bottom of the Quarto
document and create a new header called `## Filtering data`. Below that
header, create a new code chunk with {{< var keybind.chunk >}}. Then we
can write out the code together.

```{r filter-logic-condition-glykemi}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  filter(glykemi == 1)
```

We would "read" this code as:

> Take the post_meal_data dataset *and then* filter so that only rows
> where `glykemi` is equal to 1 are kept.

When a row in the `glykemi` column has the value `1`, that row is kept.
Otherwise, it is dropped. Let's try out another logical condition with
`filter()`. Let's see how many people are above "overweight" (BMI of 25)
category, which we would use `>=` for. Within the same code chunk, add
the following code:

```{r filter-logic-condition-bmi}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  filter(BMI >= 25)
```

Run this by using {{< var keybind.run >}}. We'll see that more than half
of the participants are above overweight. Let's see how many are in the
`"CTR"` group, again, within the same code chunk.

```{r filter-logic-condition-ctr}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  filter(Group == "CTR")
```

Looks like there is about half the people in the control group.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Reinforce the importance of being very careful with combining logic
conditions and to double and triple check that how you understand the
logic is correct.

For the code chunk comparing the true with the false, type this out
yourself and tell the participants to just watch. Explain as you go
along.
:::

What if we wanted to see who is overweight *and* who is in the control
group? We would use the `|` ("or") and `&` ("and") when we want to
combine conditions across columns. Be careful with these operators and
when combining logic conditions, as they can sometimes work differently
than our human brains interpret them (speaking from experience). For
`&`, both sides must be `TRUE` in order for the combination to be
`TRUE`. For `|`, only one side needs to be `TRUE` in order for the
combination to be `TRUE`. To see how they work try these:

```{r}
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```

But be careful with missing values, as they will "propagate". For
example:

```{r}
TRUE & NA
FALSE & NA
TRUE | NA
FALSE | NA
```

The result will be `NA` in all cases. This is because R doesn't know
what to do with the `NA` value, so it returns `NA` as the result. You as
the human need to figure out what to do with the `NA` values before you
use the logic condition on them.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Return to having them code along with you.
:::

So, let's see how many people are in the control group *and* have a BMI
of 25 or more. In the same code chunk, write this code below:

```{r filter-logic-condition-bmi-ctr}
#| filename: "docs/learning.qmd"
small_post_meal_data |>
  filter(Group == "CTR", BMI >= 25)
```

When we use `,` to separate our logic conditions in `filter()`, it
interprets that as using an `&`. Looks like most people who are
overweight are in the control group. Let's replace the `,` with a `|` to
see how many people are in the control group **OR** have a BMI of 25 or
more.

```{r filter-logic-condition-bmi-or-ctr}
small_post_meal_data |>
  filter(Group == "CTR" | BMI >= 25)
```

This lists both the people who are in the control as well as anyone who
is overweight.

Let's create a new data object with the renamed code we used above plus
removing the few participants with a `1` for `glykemi`. This column is
actually called hyperglycemia or impaired glucose tolerance, and is when
the level of glucose in the blood is higher than normal, but not high
enough to be diagnosed as diabetes. This is set as `0` and `1`, likely
meaning "normal" and "high". Since there are so few of them, we'll just
remove them for now. We'll also remove the `glykemi` column since we
don't need it anymore. Create a new header called
`## Make a tidier dataframe` and then create a new code chunk below that
header with {{< var keybind.chunk >}}. Then write out the code below to
create a new data object.

```{r create-new-data-object-filtered}
#| filename: "docs/learning.qmd"
tidier_post_meal_data <- small_post_meal_data |>
  rename(
    auc_glucose = auc_pg,
    auc_insulin = auc_ins
  ) |>
  filter(glykemi == 0) |>
  select(-glykemi)
tidier_post_meal_data
```

Run this code with {{< var keybind.run >}} to see the output and see
what it looks like now. Great! Before moving on, let's style our
document with {{< var keybind.styler >}}, render the document with
{{< var keybind.render >}}, and then add and commit the changes to the
Git history with {{< var keybind.git >}}. Finally, push the changes to
your GitHub repository.

::: callout-important
Recall that the original dataset doesn't change. If we want the added
variable to be saved, we must assign it to something with `<-`.
:::

## Modify or add columns

To modify an existing column or to add a new one, we can use the
function called `mutate()`. You can use `mutate()` to compute a new
column using existing columns in your dataset. You can multiply all
values in a certain column by 2, or combine columns into a new column.
Like the other functions, the first input is the dataset and the other
arguments are columns to add or modify. The follow the form:
`new_column_name = action()`.

In our data, we have a few columns we could modify or create. The column
we could create is a ratio between insulin and glucose in the blood,
which is a very rough measure of insulin sensitivity. This is calculated
by dividing the area under the curve of insulin by that of glucose.
We'll call this new column `insulin_glucose_ratio`, where a high value
means insulin insensitive (or resistant) and a low value means insulin
sensitive. Let's create a new header called `## Modifying columns` and
then create a new code chunk below that header with
{{< var keybind.chunk >}}. We'll use some math with the `/` symbol to
divide values.

```{r create-column}
#| filename: "docs/learning.qmd"
tidier_post_meal_data |>
  mutate(insulin_glucose_ratio = auc_insulin / auc_glucose)
```

Run the code with {{< var keybind.run >}} to see the new column.

The other column we could *modify*, rather than create, is the `Group`
column. This is a character column with abbreviations rather than full
names as values. We can continue using the same `mutate()` by adding a
comma to the end of the previous line and then adding the new column.
For this, we will need to use the `if_else()` function to change the
values. The `if_else()` function takes three arguments: the condition,
the value if the condition is `TRUE`, and the value if the condition is
`FALSE`. The @tbl-logic-operators is useful to use as a reference to
help create the logic condition. So if the `Group` column is equal to
`"CTR"`, we want to change it to `"Control"`, otherwise change it to
`"First-degree relative"`.

```{r modify-column}
#| filename: "docs/learning.qmd"
tidier_post_meal_data |>
  mutate(
    insulin_glucose_ratio = auc_insulin / auc_glucose,
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  )
```

Run this code with {{< var keybind.run >}} to see the new column. To
make use of this new data object, we should assign it to a new object.
So add `<-` and a new object name to the beginning of the code:

```{r assign-modified-columns-to-object}
#| filename: "docs/learning.qmd"
tidied_post_meal_data <- tidier_post_meal_data |>
  mutate(
    insulin_glucose_ratio = auc_insulin / auc_glucose,
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  )
```

Run this code with {{< var keybind.run >}} to see the new column. We'll
use this new data object in our next session.

Before moving on, run `{styler}` with {{< var keybind.styler >}}, render
with {{< var keybind.render >}}, and then add and commit all the files
changed to the Git history using {{< var keybind.git >}} before pushing
to GitHub.

## :technologist: Exercise: Practice with filter and mutate

**Time: \~15 minutes.**

Time to practice using these two functions! First, create a new header
called `## Practice with filter and mutate` at the bottom of the Quarto
document. Then create a new code chunk below that header with
{{< var keybind.chunk >}}. Copy and paste the code template below into
the code chunk before doing the tasks. Then complete the tasks by
replacing the `___` with what is asked in the task.

``` {.r filename="docs/learning.qmd"}
# Task 1.
post_meal_data |>
  ___(___ >= ___ & ___ <= ___ & ___ == ___)

# Task 2.
post_meal_data |>
  ___(
    ___ = (___ / ___) * 100
  ) |> 
  ___(___ > ___)
```

1.  Use `filter()` to find how many participants have a BMI of more than
    or equal to 20 *and* less than or equal to 40, *and* are in the
    "FDR" `Group`. How many are there?

2.  Create a `percent_body_fat` column with `mutate()` by dividing the
    `Fat.mass.DXA` by the `Weight`, then multiplying by 100. *Hint*: In
    R, use `*` to multiply and `/` to divide. Then pipe to `filter()` to
    find out how many participants have a `percent_body_fat` greater
    than 30.

Once done, style the document with {{< var keybind.styler >}}, render
the document with {{< var keybind.render >}}, and add and commit these
changes to the Git history with {{< var keybind.git >}}. Finally, push
to GitHub to synchronize with your GitHub repository.

```{r solution-pipe-filter-mutate}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# Task 1.
post_meal_data |>
  filter(BMI >= 25 & BMI <= 40 & Group == "FDR")

# Task 2.
post_meal_data |>
  mutate(
    percent_body_fat = (Fat.mass.DXA / Weight) * 100
  ) |>
  filter(percent_body_fat > 30)
```

{{< text_snippet sticky_up >}}

## Summary

-   Use the `select()` function to select specific columns in a dataset.
-   Use the `rename()` function to rename columns in a dataset.
-   Use the `filter()` function to filter rows based on a logical
    condition.
-   Use the `mutate()` function to modify existing or add new columns in
    a dataset.
-   Use the pipe `|>` to write easy-to-read code, similar to reading a
    text consisting of multiple sentences.

{{< include /includes/_code-appendix.qmd >}}
