# Wrangling with visualizing {#sec-wrangling-with-visualizing}

```{r setup}
#| include: false
library(tidyverse)
post_meal_data <- read_csv2(here::here("data/post-meal-insulin.csv"))
tidied_post_meal_data <- post_meal_data |>
  select(id, Age, BMI, Group, auc_pg, auc_ins, glykemi) |>
  rename(
    auc_glucose = auc_pg,
    auc_insulin = auc_ins
  ) |>
  filter(glykemi == 0) |>
  select(-glykemi) |>
  mutate(
    insulin_glucose_ratio = auc_insulin / auc_glucose,
    Group = if_else(Group == "CTR", "Control", "First-degree relative")
  )
```

<!-- TODO: Add an introduction here -->

## Learning objectives

This session's **overall learning outcome** is to:

1.  Combine the wrangling tools with visualizing to continue exploring
    the data, while recognizing the role that colours play in making
    your figures accessible.

**Specific objectives** are to:

1.  Apply what was learned separately in the wrangling and visualizing
    sessions by piping the output of one into the other.
2.  Recognize and explain why the choice of colour matters when it comes
    to having figures be accessible and understandable to more people,
    especially to those with colour blindness.
3.  Use features of `{ggplot2}` to make plots that continue three or
    more variables in them, through the use of colours and facetting.

## Piping your wrangling into plots

In the previous session, we learned how to wrangle data using the
`{dplyr}` package. In this session, we're going to show and highlight
how useful and powerful it is to do some wrangling and directly pipe the
output into a plot to visualize it. Often, you might need to do some
wrangling only for a specific plot, and it isn't necessary to save the
wrangled data as an intermediate object. So instead, we'll immediately
pipe the wrangled data into a plot.

Based on our data, let's say we want to see what the distribution of the
`insulin_glucose_ratio` with those who are overweight or not as a
categorical variable using BMI greater than or equal to 25 as the
threshold. Since we don't have a variable for being overweight, we need
to create one. We can use the `mutate()` function to create a new column
called `overweight` with a value for being overweight if `BMI >= 25` or
less than that. Before we create the plot, we need to decide, what type
of plot do we want. Since we have a continuous and a discrete variable,
these are a few good or common ones we could use:

-   `geom_boxplot()`, which makes boxplots that show the median and a
    measure of range in the data. Boxplots are generally pretty good at
    showing the spread of data. However, like the reading task about
    "bar plots", boxplots can still hide your actual data. It is
    generally fine to use, but a better geom might be jitter or violin.
-   `geom_jitter()`, which makes a type of scatterplot, but for discrete
    and continuous variables. A useful argument to `geom_jitter()` is
    `width`, which controls how wide the jittered points span from the
    center line. This plot is much better than the boxplot since it
    shows the *actual data*, and not summaries like a boxplot does.
    However, it is not very good when you have lots of data points.
-   `geom_violin()`, which shows a density distribution like
    `geom_density()`. This geom is great when there is a lot of data and
    `geom_jitter()` may otherwise appear as a mass of dots.

The way you use any of these geoms is the same. If you use one, you can
use another. So we'll show how to use `geom_jitter()`, because our data
are quite small.

At the bottom of the Quarto document, create a new section header called
`## Visualizing with a jitter plot`. Below that header, create a new
code chunk with {{< var keybind.chunk >}}.

```{r discrete-continuous-variables}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  mutate(overweight = if_else(BMI >= 25, "Overweight", "Not overweight")) |>
  ggplot(aes(x = overweight, y = insulin_glucose_ratio)) +
  geom_jitter()
```

Run this code chunk with {{< var keybind.run-code >}} and you'll see a
plot with a bunch of points on them. Looking at the plot, we can see
that there are more participants that are insulin insensitive
(resistant) in the overweight group than the non-overweight group. We
can modify the width of the jittered points by adding the argument
`width` to `geom_jitter()`. For now, we won't add it.

## Plotting a third or more variables

There are many ways to visualize additional variables in a plot and
further explore your data. For that, we can use `{ggplot2}`'s colour,
shape, size, transparency ("alpha"), and fill aesthetics, as well as
"facets". Faceting in `{ggplot2}` is a way of splitting the plot up into
multiple plots when the underlying aesthetics are the same or similar.

The most common and "prettiest" way of adding a third variable is by
using colour. Let's check the relationship between `Age` and
`auc_insulin` in a scatterplot with `geom_point()`, while using `Group`
in colour as the third variable.

First, create a new header called `## Plotting three or more variables`
and a new code chunk with {{< var keybind.chunk >}}. Then, write out the
code below:

```{r using-colour}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  filter(BMI < 30) |>
  ggplot(aes(x = Age, y = auc_insulin, colour = Group)) +
  geom_point()
```

Run this with {{< var keybind.run-code >}}. One thing you'll quickly
notice is that there is a clear grouping in ages between those above and
below 40. So let's make a new column called `young` that groups people
into two categories of being above or below 40 years of age. And instead
of looking at the relationship between `Age` and `auc_insulin`, let's
look at the relationship between `BMI` and `insulin_glucose_ratio`.

With piping, it is really easy to insert a wrangling step in the
sequence. So we'll insert a `mutate()` below the `filter()` and make use
of the `if_else()` function to create the new column `young`. After
that, we would like to split the plots up into two subplots, one for
each age group. We can do this using `facet_grid()`.

`facet_grid()` splits the plot up into multiple subplots. For faceting
to work, at least one of the first two arguments to `facet_grid()` is
needed. The first two arguments are:

-   `cols`: The discrete variable to use to facet the plot column-wise
    (i.e. side-by-side).
-   `rows`: The discrete variable to use to facet the plot row-wise
    (i.e. stacked on top of each other).

For both `cols` and `rows`, the variable must be wrapped by `vars()`
(e.g. `vars(young)`). Using the existing code we have, modify it so it
looks like the code below:

```{r using-faceting-cols}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  filter(BMI < 30) |>
  mutate(young = if_else(Age < 40, "Young", "Old")) |>
  ggplot(aes(x = BMI, y = insulin_glucose_ratio, colour = Group)) +
  geom_point() +
  facet_grid(cols = vars(young))
```

Run this with {{< var keybind.run-code >}}. You'll see two plots
side-by-side, one for each age group. Super neat! Right away, we can see
that there are more older people who have first-degree relatives with
diabetes, which makes sense since older people will likely have older
relatives, and age is a big factor in diabetes. You can also facet by
`rows`, by replacing the `cols` argument:

```{r using-faceting-rows}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  filter(BMI < 30) |>
  mutate(young = if_else(Age < 40, "Young", "Old")) |>
  ggplot(aes(x = BMI, y = insulin_glucose_ratio, colour = Group)) +
  geom_point() +
  facet_grid(rows = vars(young))
```

Which do you find more informative?

We can modify the axis labels to have clearer descriptions by using the function `labs()`.
To change the y-axis title, use the `y` argument in `labs()`.
For the x-axis, it is `x`. Let's add x and y axis labels to our existing plot.

```{r labels}
#| filename: "docs/learning.qmd"
tidied_post_meal_data |>
  filter(BMI < 30) |>
  mutate(young = if_else(Age < 40, "Young", "Old")) |>
  ggplot(aes(x = BMI, y = insulin_glucose_ratio, colour = Group)) +
  geom_point() +
  facet_grid(rows = vars(young)) +
  labs(
    x = "Body mass index (BMI)",
    y = "Insulin to glucose ratio"
  )
```

When we run that with {{< var keybind.run-code >}}, we now have labels that are more descriptive!

::: callout-tip
To save the plot you created, you can use the `ggsave()` function.
:::

Before moving on, run `{styler}` with {{< var keybind.styler >}}, render
the Quarto document with {{< var keybind.render >}}, and commit the
changes to the Git history with {{< var keybind.git >}}. Lastly, push
the changes to GitHub.

## :technologist: Exercise: Creating some multi-dimensional plots

**Time: \~15 minutes.**

Time to try this out yourself! Create two figures that are side by side
like we did in @sec-visualization with the Quarto code chunk
`#| layout-ncol: 2`. Create a new header in the `docs/learning.qmd` file
called `## Make multi-dimensional plots!`, then below it copy and paste
the template code chunk shown below into your document.

````{.markdown filename="docs/learning.qmd"}
```{{r}}
#| label: fig-multidimensional
#| fig-cap: "Two, mult-dimensional plots of our data."
#| fig-subcap:
#|   - "Distribution of percent body fat by group and age category."
#|   - "Relationship between BMI and bone mineral density, split by insulin sensitivity."
#| layout-ncol: 2
# Task 1.
post_meal_data |>
  mutate(
    ___ = (___ / ___) * 100,
    ___ = if_else(___ < 40, ___, ___)
  ) |>
  ___(___(x = ___, y = ___, colour = ___)) +
  ___(width = 0.2)

# Task 2.
post_meal_data |>
  mutate(
    insulin_glucose_ratio = (auc_ins / auc_pg),
    insulin_insensitive = if_else(___ > ___(___), ___, ___)
  ) |>
  ___(___(x = ___, y = ___)) +
  ___() +
  ___() +
  ___(cols = vars(___))
```
````

1.  Show the distribution of `percent_body_fat` by `Group` and `Age`
    category following these steps:
    1.  Create a new column called `percent_body_fat` that calculates
        the percentage of body fat by dividing `Fat.mass.DXA` by
        `Weight` and multiplying by 100. You wrote this code previously
        in another exercise, you can re-use that code if you would like.
    2.  Create a new column called `young` that groups people into two
        categories of being above (`"Old"`) or below (`"Young"`) 40
        years of age by using `if_else()`.
    3.  Use `ggplot()` and `aes()` to create the plot by putting `Group`
        on the x-axis, `percent_body_fat` on the y-axis, and colour the
        points by `young`. Use `geom_jitter()` to show the data points.
        The argument `width = 0.2` has already been set for you to have
        the points cluster together a bit more
2.  Show the relationship between BMI and bone mineral density,
    splitting it by insulin sensitivity by following these steps:
    1.  A new column called `insulin_glucose_ratio` has already been
        included in the template for you to use. Below this code, create
        a new column called `insulin_insensitive` that groups people
        into two categories of being above (`"Insensitive"`) or below
        (`"Sensitive"`) the `mean()` of `insulin_glucose_ratio` by using
        `if_else()`.
    2.  Use `ggplot()` and `aes()` to create the plot by putting `BMI`
        on the x-axis, `Bone.mineral.DXA` on the y-axis, and colour the
        points by `insulin_insensitive`.
    3.  Use `geom_point()` to show the data points, then also use
        `geom_smooth()` to show a line of best fit.
    4.  Use `facet_grid()` to split the plot by `insulin_insensitive`.

When you complete all the tasks, run `{styler}` using
{{< var keybind.styler >}}, then render the document with
{{< var keybind.render >}} to make sure it runs and to see the output.
End with adding and committing the changes to the Git history with
{{< var keybind.git >}} and pushing the changes to GitHub.

```{r solution-multidimensional}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# This is an example of what the code will look like:
#| label: fig-multidimensional
#| fig-cap: "Two, mult-dimensional plots of our data."
#| fig-subcap:
#|   - "Distribution of percent body fat by group and age category."
#|   - "Relationship between BMI and bone mineral density, split by insulin sensitivity."
#| layout-ncol: 2
# Task 1.
post_meal_data |>
  mutate(
    percent_body_fat = (Fat.mass.DXA / Weight) * 100,
    young = if_else(Age < 40, "Young", "Old")
  ) |>
  ggplot(aes(x = Group, y = percent_body_fat, colour = young)) +
  geom_jitter(width = 0.2)

# Task 2.
post_meal_data |>
  mutate(
    insulin_glucose_ratio = (auc_ins / auc_pg),
    insulin_insensitive = if_else(insulin_glucose_ratio > mean(insulin_glucose_ratio), "Insensitive", "Sensitive")
  ) |>
  ggplot(aes(x = BMI, y = Bone.mineral.DXA)) +
  geom_point() +
  geom_smooth() +
  facet_grid(cols = vars(insulin_insensitive))
```

{{< text_snippet sticky_up >}}

## :book: Reading task: Using accessible colours in your plot

**Time: \~5 minutes**

Colour blindness is common in the general population, with red-green
colour blindness affecting about 8% of men and 0.5% of women. To make
your graph more accessible to people with colour blindness, you need to
consider the colours you use. For more detail on how colours look to
those with colour blindness, check out this
[documentation](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)
from the `{viridis}` package.

The `{viridis}` colour scheme (also developed as an R package) was
specifically designed to represent data to all colour visions (including
as a grayscale, e.g. from black to white). There is a really informative
[talk on YouTube](https://www.youtube.com/watch?v=xAoljeRJ3lU) on this
topic.

When using colours, think about what you are trying to convey in your
figure and how your choice of colours will be interpreted. You can use
built-in colour schemes or create your own. For now, let's stick to
using built-in ones. There are two we can start with: the `{viridis}`
and the [ColorBrewer](https://colorbrewer2.org/) colour schemes. Both
are well designed and are colour blind friendly. For this workshop, we
will only cover the `{viridis}` package.

::: callout-tip
You can modify ggplot2 colour schemes using many other pre-defined
palettes by installing new R packages, including scientific journal
colour palettes ([ggsci](https://nanx.me/ggsci/index.html)) and even a
Wes Anderson ([wesanderson](https://github.com/karthik/wesanderson)) or
a Stubio Ghibli ([ghibli](https://ewenme.github.io/ghibli/)) colour
palette!
:::

Continue to the exercise once you've finished reading this section to
test out the `{viridis}` colour scheme.

## :technologist: Exercise: Changing the colour schemes

**Time: \~8 minutes.**

Change the colour schemes on some plots. Start with a base plot object
to work from that has two discrete variables. Create a new Markdown
header called `## Changing colours` and create a new code chunk below it
using {{< var keybind.chunk >}}. Copy and paste the code below into the
new code chunk.

``` r
# Tasks 1. and 2.
post_meal_data |>
  mutate(hyperglycemia = if_else(glykemi == 1, "Yes", "No")) |>
  ggplot(aes(x = Group, fill = hyperglycemia)) +
  geom_bar(position = position_dodge()) +
  ___()

# Tasks 3. and 4.
post_meal_data |>
  ggplot(aes(x = auc_pg, y = auc_ins, colour = BMI)) +
  geom_point() +
  ___()
```

::: callout-tip
If you need help, use the `help()` or `?` functions in RStudio to look
over the documentation for more information for the functions we get you
to use below.
:::

1.  Change the colour to the `{viridis}` scheme by using the
    `scale_fill_viridis_d()` function, adding it to the end of the
    `# Task 1. and 2.` code. Because the variables are discrete, the
    function ends with `_d`.

2.  `{viridis}` has several palettes. Add the argument
    `option = "magma"` to the `scale_fill_viridis_d()` function. Run the
    function again and see how the colour changes. Then, change
    `"magma"` to `"cividis"`.

3.  Now, let's practice using the colour schemes on a plot with
    continuous variables. For the `# Tasks 3. and 4.` code, use
    `scale_colour_viridis_c()` at the end of the `ggplot()` function. We
    use `_colour_` instead of `_fill_` in the `scale_` name because,
    well, we are using the `colour =` argument. The `_c` at the end
    indicates the variable are continuous.

4.  Similar to task 2 above, use the `option = "inferno"` argument to set the
    palette and see how the colour changes. Select which
    colour scheme you would like.

Once you are done, style your code with `{styler}` using
{{< var keybind.styler >}}, render the document with
{{< var keybind.render >}}, and commit the changes to the Git history
with {{< var keybind.git >}}, followed by pushing to GitHub.

```{r solution-colour-schemes}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# Tasks 1. and 2.
post_meal_data |>
  mutate(hyperglycemia = if_else(glykemi == 1, "Yes", "No")) |>
  ggplot(aes(x = Group, fill = hyperglycemia)) +
  geom_bar(position = position_dodge()) +
  scale_fill_viridis_d()
# scale_fill_viridis_d(option = "magma")

# Tasks 3. and 4.
post_meal_data |>
  ggplot(aes(x = auc_pg, y = auc_ins, colour = BMI)) +
  geom_point() +
  scale_colour_viridis_c()
# scale_colour_viridis_c(option = "inferno")
```

{{< text_snippet sticky_up >}}

## Summary

-   Pipe `|>` the output of your wrangling directly into making a plot
    to streamline your code.
-   Use `geom_jitter()`, `geom_boxplot()`, or `geom_violin()` to show the
    relationship between a continuous and a discrete variable.
-   To plot more dimensions, use colour, the X axis, the Y axis, or
    `facet_grid()`.
-   Use colour blind-friendly palettes, such as `{viridis}`.

## Survey

Please complete the survey for this session:

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
