---
execute:
  eval: false
---

# Setting up Git and GitHub {#sec-setup-git-and-github}

After having read about Git during the reading task in @sec-pre-reading,
you now need to setup Git to be ready for the workshop. In order to use
Git properly, you need to set some options that tell Git your name and
email address so that Git can use that information when recording
changes you make to your files. You'll learn more about that and how it
looks like during the workshop. For now, you just need to run one of the
r3 helper functions to help setup Git for you. In your RStudio, go to
the Console and type (or copy and paste) the following function:

```{r setup-git-config}
#| filename: "Console"
r3::setup_git_config()
```

Hit enter and follow the instructions. Finally, type and run this next
function to make sure everything is working with your setup. When you
complete the survey later, you will need to copy and paste the output of
this function. For now, run it to confirm everything is setup properly.

```{r run-check-setup}
#| filename: "Console"
r3::check_setup()
```

If everything is fine, you should see something that looks like:

```
Checking R version:
✔ Your R is at the latest version of 4.4.3!
Checking RStudio version:
✔ Your RStudio is at the latest version of 2025.05.1.513!
Checking Git config settings:
✔ Your Git configuration is all setup!
  Git now knows that:
  - Your name is 'Luke W. Johnston'
  - Your email is 'lwjohnst@gmail.com'
```

After you are done, you need to [create a GitHub
account](https://github.com/join). Remember or record your username, as
we will ask you for it in the pre-workshop survey. Make sure to
**remember your password**, ideally saving it in your [password
manager](https://en.wikipedia.org/wiki/Password_manager) if you
installed one already or during the pre-workshop tasks so that it can
store your password for you.

::: {.callout-note appearance="default"}
*GitHub* is a company and website (like Google), while *Git* is a
software (like R). There is sometimes confusion about these two things
since they both say "Git". It's important to distinguish that they are
two *separate* things.
:::
