---
execute: 
  echo: false
  eval: true
---

# Schedule {#sec-schedule}

The course is structured as a series of participatory live-coding
sessions interspersed with hands-on exercises and group work, using
either a practice dataset or some other real-world dataset. There are
some lectures given, mainly at the start and end of the course. The
general schedule outline is shown in the below table. This is not a
fixed schedule of the timings of each session --- some may be shorter
and others may be longer. Instead, it is meant to be an approximate
guide and overview.

::: panel-tabset
```{r}
library(dplyr)
add_link <- function(text, icon, link) {
  colour <- "#138d75"
  added_link <- paste0("<i class='fa-solid fa-", icon, "' style='color:", colour, "'></i>&nbsp;&nbsp;<a href=\"", link, "\">", text, "</a>")
  if (link == "") {
    added_link <- paste0("<i class='fa-solid fa-", icon, "' style='color:", colour, "'></i>&nbsp;&nbsp;", text)
  }
  return(htmltools::HTML(added_link))
}

schedule <- here::here("preamble/schedule.csv") %>%
  readr::read_csv(col_types = "ccccc") %>%
  mutate(across(everything(), ~ if_else(is.na(.x), "", .x))) %>%
  mutate(
    `Session topic` = list(`Session topic`, icon, link) %>%
      purrr::pmap(~ add_link(..1, ..2, ..3)) %>%
      unlist()
  ) %>%
  select(-icon, -link)

list_day_schedule <- function(data, day) {
  data %>%
    filter(Day == day) %>%
    select(-Day) %>%
    knitr::kable()
}
```

## Day 1 {.unnumbered}

```{r}
list_day_schedule(schedule, "Day 1")
```

## Day 2 {.unnumbered}

```{r}
list_day_schedule(schedule, "Day 2")
```

## Day 3 {.unnumbered}

```{r}
list_day_schedule(schedule, "Day 3")
```
:::
