---
execute:
  eval: false
---

# Extras: Arranging rows {#sec-extra-arrange}

You may want to sort your rows by a specific column so that values are
arranged in ascending or descending order. This can be done using the
function called `arrange()`. Again, `arrange()` takes the dataset as the
first argument, followed by the columns that you wish to arrange data
by. By default, `arrange()` orders in *ascending* order.

```{r}
#| filename: "docs/learning.qmd"
post_meal_data |>
  arrange(Age)
```

`arrange()` also arranges parameters of type `character` alphabetically:

```{r}
#| filename: "docs/learning.qmd"
post_meal_data |>
  arrange(Group)
```

If we want to order the column based on descending order, this can be
done with `desc()`.

```{r}
#| filename: "docs/learning.qmd"
post_meal_data |>
  arrange(desc(Age))
```

You can also order your data by multiple columns. For instance, we could
arrange first by `Group` and then by `age`.

```{r}
#| filename: "docs/learning.qmd"
post_meal_data |>
  arrange(Group, Age)
```
